---
title: "Quantifying the human influence on drought"
author: "Worked example 10.1"
---

```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
library(hydroDrought)
# Load the data
guadiana <- read_tsv("../../data/Guadiana_Delay_LogEff_including_Psim.txt") %>%
  dplyr::select(time = "date.yyyymmdd",  Qsim, Qobs)  
# append_group(by = "day")
```

In this worked example, the catchment-scale observation-based method is used to quantify human influence on hydrological drought, with the threshold level method for drought calculation (Section 10.4.1; Fig. 10.5). We compare two time series from the Upper-Guadiana catchment, one with and one without human influence. The time series without human influence can be calculated from a paired catchment analysis, an upstream-downstream comparison, observed-naturalised comparison, or pre-post disturbance analysis (Section 10.4.1; Table 10.1). In this example, we use the observed-naturalised approach (Fig. 10.6).

# Loading the data

We begin by loading the two time series from the Upper-Guadiana dataset: the benchmark (naturalised) time series and the human-influenced time series. The benchmark time series is simulated, as described in Section 9.3.3. In this case, the observed time series from the Upper-Guadiana catchment is the human-influenced time series (Section 4.5.X).

```{r}
library(tidyverse)
library(lubridate)
library(hydroDrought)

### Filter to the period 1960-2000 
### Rename the columns
### Calculate the water year
guadiana_full <- guadiana %>%
	dplyr::select(time, benchmark = Qsim, influenced = Qobs) %>%
  	mutate(
    	year = water_year(time)
	) %>%
	filter(year >= 1960,  year <= 2000) 

print(guadiana_full)
range(guadiana_full$time)
```

```{r, echo=FALSE}
fig.cap <- "Figure 10.xxxx: Time series showing the benchmark and human-influenced time series. Discharge is plotted on a log scale to emphasize low flow behavior. The grey region shows the comparison period for human influence (1981-2000)."

plot_df <- guadiana_full %>%	
	mutate(jdate = yday(time)) %>% 
	#dplyr::select(-time) %>% 
	pivot_longer(cols = c("benchmark", "influenced"))	 %>%
	rename(facet = name) %>%
	mutate(line = facet) %>%
	mutate(line = factor(line, levels = c("thresh_basis", "benchmark", "influenced"), labels = c("Threshold\nBasis", "Benchmark", "Human\nInfluenced"))) %>%
	mutate(facet = factor(facet, levels = c("thresh_basis", "benchmark", "influenced"), labels = c("Threshold Basis", "Benchmark", "Influenced"))) %>%
	mutate(label = "Threshold")

```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot the 1960-2000 timee series
p <- ggplot(plot_df, aes(x=time)) %>%
	 + geom_rect(aes(xmin=as.Date("1981-01-01"), xmax= as.Date("2000-12-31"), ymin=0, ymax=Inf), fill = "grey90", alpha = 0.2) %>% 
	 + geom_line(aes( y=value, colour = line)) %>%
	 #+ geom_vline(xintercept = c(as.Date("1981-01-01"), as.Date("2000-12-31")), linetype = "dashed", size = 0.5, colour = "grey30") %>%
	+ scale_colour_manual(name = "", values = c("black", "#377eb8"), limits = c("Benchmark", "Human\nInfluenced")) %>%
	+ facet_grid(facet~.) %>%
	+  scale_y_log10(name = "Discharge (mm/day)") %>%
	+ annotation_logticks(sides = "lr") %>%
	+ theme(panel.grid.minor = element_blank())	 

### Display figure (stored in variable p)
p
```

# Threshold calculation

We use a daily varying threshold, the Q~80~, calculated based on a smoothed 30-day moving average series over the reference period (1960-2000). The threshold is derived from the benchmark time series.

```{r}
### First apply a 30-day moving average smoother
### Then compute the Q80 for each day of the year
threshold <- guadiana_full %>%
  mutate(discharge = moving_average(benchmark, n = 30, sides = "center")) %>%
  var_threshold(vary.by = "day", fun = lfquantile, exc.freq = 0.80) 

print(threshold)
```



```{r, echo=FALSE}
fig.cap <- "Figure 10.xxxx: Benchmark discharge for the period 1960-2000 showing seasonality, with each line representing a year. The smoothed Q~80~ threshold is shown in bold red."

### Add julian dates and a plot date
plot_df <- guadiana_full %>%
	mutate(jdate = yday(time)) %>%
	mutate(plot_date = as.Date(paste0("2000-01-01"))+jdate - 1)

plot_thresh <- threshold %>% 
	mutate(jdate = yday(day)) %>%
	mutate(plot_date = as.Date(paste0("2000-01-01"))+jdate - 1)
```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot the threshold against flows
p <- ggplot(plot_df, aes(x=plot_date)) + 
	geom_line(aes(group=year, y=benchmark), alpha = 0.2) + 
	geom_line(data = plot_thresh, aes(y=threshold), colour="red", size = 1) + 
	scale_x_date(name = , "Month", date_labels = "%b") +
	scale_y_continuous(name = "Discharge (mm/day)")

### Display figure (stored in variable p)
p
```

# Calculate the benchmark drought characteristics (drought duration and deficit volume)

Drought events in the benchmark time series are then identified with the threshold level method (Section 5.4.1) for a selected period. Note that this time period can differ from the reference period used to calculate the benchmark threshold. In this example, we derive drought events over the period 1981-2000, which equals the period with the main human intervention in the Upper Guadiana catchment. To limit the problem of dependent droughts, consecutive drought events with an inter-event time less than or equal to 10 days are pooled into a single drought event (Worked Example 5.5). To remove minor droughts, only drought events with a duration of more than 10 days are kept (Section 5.x). Average drought duration and deficit volume are then calculated for the resulting series of events. We find for the naturalised benchmark time series a mean duration of 91 days and a mean deficit volume of 0.67 mm.

```{r}
# initialize empty list for events
events <- list(benchmark = NULL, influenced = NULL)

# initialize empty list for final drought characteristics
drought.char <- list(benchmark = NULL, influenced = NULL) 

# function that computes the drought characteristics given a table of events
summarize_dc <- function(x) {
  c("mean.duration" = as.double(mean(x$duration)), 
    "mean.deficit" = mean(x$volume))
}
```

```{r}

### The drought_events function assumes all flows are in volume per second, so we must convert the threshold from vol/day
threshold_mm_s <- threshold %>%
	mutate(threshold = threshold/(24*60*60))

### Calculate the drought events for the benchmark time series
### Filter to 1981-2000
### use the benchmark column as discharge
### The drought_events function assumes all flows are in volume per second, so we must convert from vol/day
events$benchmark <- guadiana_full %>%
  filter(year >= 1981, year <= 2000) %>%
  #rename(discharge = benchmark) %>%
  mutate(discharge = benchmark/(24*60*60)) %>%
  drought_events(
    threshold = threshold_mm_s,
    pooling = "inter-event",
    pooling.pars = list(min.duration = 10, min.vol.ratio = Inf)
  )  %>%
  filter(duration > 10)

# calculate the drought characteristics for the benchmark time series
drought.char$benchmark <- summarize_dc(events$benchmark)

print(events$benchmark)
```

# Calculate the human-influenced drought characteristics

Similarly, drought events in the human-influenced time series are identified. Note that we are using the same period (1981-2000) as was used to derive drought indices for the benchmark time series (Step 3). Furthermore, we are applying the same inter-event criterion and exclusion of minor drought events (Step 3). We find for the human-influenced time series a mean duration of 304 days and a mean deficit volume 3.59 mm.


```{r}
### Calculate the drought events for the human-influenced time series
### Filter to 1981-2000
### use the influenced column as discharge
events$influenced <- guadiana_full %>%
  filter(year >= 1981 & year <= 2000) %>%
  # rename(discharge = influenced) %>% 
  mutate(discharge = influenced/(24*60*60)) %>%
  drought_events(threshold = threshold_mm_s,
                 pooling = "inter-event",
                 pooling.pars = list(min.duration = 10, min.vol.ratio = Inf))  %>%
  filter(duration > 10)

# calculate the drought characteristics for the human influenced time series
drought.char$influenced <- summarize_dc(events$influenced)
```

```{r, echo=FALSE}
fig.cap <- "Figure 10.xxxx: Benchmark (top) and human-influenced (bottom) flows during the comparison period 1981-2000. The smoothed Q80 threshold is shown in red. Discharge is plotted on a log scale to emphasize low flow behavior."

### Create a dataframe of the threshold for each day
thresh_plot <- guadiana_full %>%
	dplyr::select(time, year) %>%
	mutate(jdate = yday(time))

plot_thresh <- thresh_plot %>%
	mutate(line = "threshold", facet = "benchmark") %>%
	bind_rows(thresh_plot %>% mutate(line = "threshold", facet = "influenced")) %>%
	left_join(threshold %>% mutate(jdate = yday(day)) %>% dplyr::select(-day), by = "jdate") %>%
	mutate(source = line) %>%
	rename(discharge = threshold)
	
### Merge discharge with the threshold and create columns for plotting
plot_df <- guadiana_full %>%
	pivot_longer(cols = c("benchmark", "influenced"), names_to = "source", values_to  = "discharge") %>%
	mutate(jdate = yday(time), year = year (time)) %>%
	mutate(line = source) %>%
	mutate(facet = source) %>%
	bind_rows(plot_thresh) %>%
	mutate(line = factor(line, levels = c("benchmark", "influenced", "threshold"), labels = c("Benchmark", "Human\nInfluenced", "Threshold"))) %>%
	mutate(facet = factor(facet, levels = c("benchmark", "influenced", "threshold"), labels = c("Benchmark", "Human\nInfluenced", "Threshold"))) %>%
	mutate(plot_date = as.Date(paste0("2000-01-01"))+jdate - 1)
```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot flow with time relative to the threshold
p <- ggplot(plot_df, aes(x=time)) %>%
	+ geom_line(aes( y=discharge, colour = line)) %>%
	+ facet_grid(facet~.) %>%
	+ scale_colour_manual(name = "", values = c("black", "#377eb8", "red"), limits = c("Benchmark", "Human\nInfluenced", "Threshold")) %>%
	+ scale_x_date(name = "Date", date_labels = "%Y") %>%
	+  scale_y_log10(name = "Discharge (mm/day)") %>%
	+ annotation_logticks(sides = "lr") %>%
	+ theme(panel.grid.minor = element_blank())	%>%
	+ coord_cartesian(xlim=c(as.Date("1981-01-01"), as.Date("2001-01-01"))) %>%
	+ theme(legend.position="bottom")

### Display figure (stored in variable p)
p
```



# Comparison of drought characteristics    
In this final step, we compare the drought characteristics between the benchmark and human-influenced time series by calculating the percentage difference between the two series, $\Delta DI$ :

$$\Delta DI = \frac{DI_{HI} - DI_{BM}}{DI_{BM}} \cdot 100$$

where $\Delta DI$ is the percentage difference in drought index (DI) between the human-influenced ($DI_{HI}$) and benchmark ($DI_{BM}$) time series.

```{r}
drought.char
```
```{r}
(drought.char$influenced - drought.char$benchmark) / drought.char$benchmark * 100
```

We find that the difference in mean duration is 236% and in mean deficit volume 434%. To conclude, streamflow droughts in the Upper-Guadiana catchment have increased in duration and mean deficit, with on average 236% and 434%, respectively, in the human-influenced situation compared to the naturalised situation.


