---
title: "Drought Impact Model"
author: "Worked example 12.2"
---


```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

In this example, we will determine how agricultural drought impacts in Germany are related to drought severity by defining fitting an impact model to data from the past and by using it to make predictions of impact likelihood. 

# Loading the data

We begin by loading the agricultural drought impact data, collected annually, and the drought indices, which are all stored in the package hydroDrought. This data are loaded when the hydroDrought package is loaded. The SPI and SPEI data are provided at a monthly resolution, so we must reduce this to an annual resolution to match the impact series. We choose to filter only to July, based on our prior knowledge of agricultural practices. 

```{r}
library(tidyverse)
library(hydroDrought)

## Push these into the function
library(gridExtra)
library(grid)

spi_de <- spi_de %>%
	filter(month == 7)

### Merge dataframes using a full join (i.e. all records are included, even if data is missing)
spi_impacts <- spi_de %>%
	full_join(impacts_de, by = "year")  %>%	### Join on year
	drop_na(impact)	%>%						### Drop rows with NA in the impact column
	print(spi_impacts)						### Check results
```
# Preliminary visual comparison
As a preliminary comparison, we can plot the time series of July SPEI-3 (in blue) overlaid on the occurrence of annual agricultural impacts (vertical red bars). 


```{r, echo=FALSE}
fig.cap <- "Figure 12.10 Time series of July SPEI-3 (in blue) plotted together with the occurrence of annual agricultural drought impacts in Germany (vertical red bars)."

### Only plot SPEI 3
plot_df <- spi_impacts %>%
    filter(index == "spei_3")

### Extract the impacts for separate plotting
plot_impacts <- plot_df %>%
    dplyr::select(year, impact) %>%
    filter(impact == TRUE) %>%
    dplyr::mutate(left = year - 0.5, right = year + 0.5)

### Create the time series plot
p <- ggplot(plot_df, aes(x=year)) %>%
    + geom_hline(yintercept=0,  color = "grey70") %>%  ### Make a horizontal axis at y= 0
    + geom_rect(data = plot_impacts, aes(xmin=left, xmax=right, ymin=-Inf, ymax=+Inf), fill='pink', alpha=0.9) %>%   ### Create a pink region around the impact
    + geom_vline(data = plot_impacts, aes(xintercept = year), linetype="dotted", color = "red") %>%   ### Add a vertical line at the impact
    + geom_line( aes(y = index_val), colour="#377eb8") %>%  ### Draw the time series of SPI12
    + theme_classic(10) %>%
    + scale_x_continuous(name = "Year", breaks = seq(1950,2100,5)) %>%
    + scale_y_continuous(name = "SPEI-3 (July)", breaks = seq(-3,3,0.5)) 
```

```{r,  echo=FALSE, fig.cap=fig.cap}
p
```
When SPEI-3 is very negative (approximately -2) in 1975 an impact occurs, but when SPEI-3 is near -1.5, we have 2 impact years (2003, 2006) and 2 non-impact years (1989, 2008). So, it appears that at a SPEI-3 of -1.5, we have an approximately 50% chance of an impact. As we approach positive SPEI-3, based on this data, the number of drought impacts drops to near 0 (there are no impact records in this database). 

We can also visualise the relationship between SPEI-3 and impact by plotting each value of SPEI-3 from the time series along the x-axis and stacking the proportion of impact vs. no impact months along the y-axis (Fig. 12.11). 

```{r, echo=FALSE}
fig.cap <- "Figure 12.11 Relationship between the proportion of years with agricultural impacts (orange field) and no impacts (green field) for: the range of (a) and (b) SPI-12 (a, c),  and (c) and (d) SPI-3 (b, d) values in during July ranging from -2 to 2."

### Create a meaningful column for impacts when plotted
plot_all_df <- spi_impacts %>%
    dplyr::mutate(impact_label = case_when( impact == TRUE ~ "Impact Reported",
        			     TRUE ~ "No Impact"
        )  ) %>%
    dplyr::mutate(impact_label = factor(impact_label, levels=c( "No Impact", "Impact Reported"))) %>%
	dplyr::mutate(index_label = factor(index, levels = c("spei_12", "spei_3", "spi_12", "spi_3"), labels = c("SPEI-12", "SPEI-3", "SPI-12", "SPI-3")))

###  Create a smoothed plot showing proportion in each category
p <- ggplot(plot_all_df, aes(x=index_val, fill = impact_label)) %>%
    + geom_density(position = "fill") %>%
    + geom_vline(xintercept=0,  color = "grey30", linetype="longdash") %>%  ### Make a vartical axis at x= 0
    + theme_classic(8) %>%
    + scale_x_continuous(name = "Drought Index in July", breaks = seq(-5,5,0.5)) %>%
    + scale_y_continuous(name = "Proportion of Years in Impact Category", labels = scales::percent_format(accuracy = 5L), expand = c(0,0)) %>%
    + scale_fill_brewer(type = "qual", palette = "Set2") %>%
    + theme(legend.position = c(0.9, 0.9), legend.title = element_blank()) %>%
    + facet_wrap(~ index_label)
```

```{r, echo=FALSE,fig.cap=fig.cap}
p
```

The relationship for SPEI3 is shown in the top-right figure (Fig. 12.11b). If we approximate the boundary between impact years (orange) and non-impact years (green), it appears there is a consistent increase in impact proportion as SPEI-3 decreases (shifts from right to left). 

SPEI-3 appears to have the most consistent relationship with impacts (Fig 12.11b), whereas the relationship with SPI-3 is confounded by some drought periods (low SPI) where the ‘No Impacts’ category dominates. If we chose a wider smoothing interval, this effect might be averaged out. For the 12-month accumulated indices (Fig 12.11a, 12.11c), years with impacts appear to be scattered throughout the SPI/SPEI range, suggesting a weaker connection. The likelihood of impacts increases for both severe drought and pluvials, further supporting the hypothesis that the 12-month indices are not useful for making the distinction between impact and no impact for this particular example using German agriculture. 

# Logistic Regression

Fig. 12.11 is a visual analogy of logistic regression, which is the more rigorous statistical method we will use to model the relationship between one or more predictors and a binary (TRUE/FALSE) predictand. Please keep this visual analogy in mind as we proceed using logistic regression to develop a quantitative impact model. Section 7.3.7 gives mMore methodological details are given in Section 7.3.7, Generalized Linear Models. 


```{r}
### Create data for fitting
fit_data <- spi_impacts %>%
    select(date, year, month, index, index_val, impact) %>%
    spread(index, index_val)

### Fit simple logistic regression using SPEI-3 as a predictor variable
### Force an intercept to be included by using + 1
spei_3_fit <- glm(impact ~ spei_3 + 1, data = fit_data, family = "binomial")
summary(spei_3_fit)

confint(spei_3_fit)
```

The fitted model appears similar to our visual analogy (Fig. 12.11), but now follows a formal logistic regression equation, with confidence intervals (Fig. 12.12). As expected from our visual analogy, the likelihood of an impact is approximately 50% for SPEI-3 of -1.1, near 100% for -2 and near 0 for positive values.  

```{r, echo=FALSE}
fig.cap <- "Figure 12.12 Relation between the drought index (SPEI-3) and predicted agricultural impact likelihood for Germany as a solid (red) line with shaded 95% confidence intervals; the distribution of SPEI-3 is shown as box plots for months with impacts (top) and months with no impacts (bottom)."
```
```{r, echo=FALSE, fig.cap=fig.cap}
### Plot the relationship with SPEI-3
logistbarplot(log.fit = spei_3_fit, log.var = "spei_3")

```

Remember from Section 7.2.3 that logistic regression converts likelihoods of a binary response into log-odds space (also called logit space). When a line fit in this space is transformed back to probability space, it produces the characteristic S-shape curve between 0 and 1. If we had used linear regression without the logit transform, we might have produced likelihoods greater than 1 or less than 0, which are not feasible. 

# Predicting drought impacts
It is now possible to make predictions based on this relationship for our fitting period or for any new values of SPI or SPEI.

```{r}
### Use the SPEI3 regression to predict the likelihood of an impact
### Keep the standard error for confidence intervals around prediction
predict_ts <- predict(spei_3_fit, newdata = fit_data, type = "link", se = TRUE)

### Add in prediction and confidence intervals
predict_df <- fit_data %>%
    mutate(predict = plogis(predict_ts$fit)) %>%
    mutate(predict_upper = plogis(predict_ts$fit + (qnorm(0.025) * predict_ts$se.fit))) %>% 
    mutate(predict_lower = plogis(predict_ts$fit - (qnorm(0.025) * predict_ts$se.fit))) %>%
    select(-spi_3, -spei_12, -spi_12) %>%
	print()
```

Making predictions based on the original training data, we can now see (Fig. 12.13) that during most years the likelihood of an agricultural impact remains below 10%. In years with low SPEI-3, the likelihood quickly peaks. In 1976, our most severe drought in this period, the predicted likelihood is approximately 90%. Not surprisingly, an impact occurs during this period. For a predicted likelihood of 50%, there is approximately the same number of years with impacts as those without, a sign of a good fit. Also, notice that the prediction intervals are not normally distributed around the estimate. They are normally distributed in logit (log-odds) space, but have been converted into impact likelihood, ensuring that neither the estimate nor the interval produces impossible likelihoods (< 0 or > 1).

```{r, echo=FALSE}
fig.cap <- "Figure 12.13 Predictions of the likelihood of an agriculture impact (blue line with grey confidence intervals) and deducted binary predicted occurrence (pink)."

### Create a time series plot
p <- ggplot(predict_df, aes(x=year)) %>%
    + geom_hline(yintercept=0.5,  color = "grey70", linetype="dotted") %>%  ### Make a horizontal axis at y= 0
   + geom_rect(data = plot_impacts, aes(xmin=left, xmax=right, ymin=-Inf, ymax=+Inf), fill='pink', alpha=0.9) %>%   ### Create a pink region around the impact
#   + geom_vline(data = plot_impacts, aes(xintercept = year), linetype="dotted", color = "red") %>%   ### Add a vertical line at the impact
    + geom_ribbon(aes(ymin = predict_lower, ymax = predict_upper), alpha = 0.2, fill = "grey20") %>%
    + geom_line( aes(y = predict), colour="#377eb8") %>%  ### Draw the time series of SPI12
    + theme_classic(8) %>%
    + scale_x_continuous(name = "Year", breaks = seq(1950,2100,5)) %>%
    + scale_y_continuous(name = "Impact Likelihood", breaks = seq(0,1,0.1), labels = scales::percent_format(accuracy = 5L), expand=c(0,0))

```
```{r, echo=FALSE, fig.cap=fig.cap}

p
```


```{r, echo=FALSE}
### Put the rest of the code here.
```



