---
title: "Worked Example 5.1: Flow Duration Curve"
---

```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Loading the Data

In this example we are going to use river flow data from the river Ngaruroro which is part of the International Data Set in the package **hydroDrought**. The dataset becomes accessible by loading the **hydroDrought** package. 
```{r}
library(tidyverse)
library(hydroDrought)
ngaruroro <- international %>%
  filter(river == "Ngaruroro") %>%
  select(data) %>%
  unnest(data) 
```

```{r, properties-dataset, echo=FALSE}
fmt_date <- function(x) trimws(format(x, format = "%e %B %Y"))
```

The complete record (`r fmt_date(min(ngaruroro$time))` to `r fmt_date(max(ngaruroro$time))`) 
of daily data from River Ngaruroro at Kuripapango (NZ) are used here to construct 
a flow duration curve (FDC) based on a daily time step, $\Delta t = 1$ day. The 
total number of $\Delta t$ intervals is $N = `r nrow(ngaruroro)`$ days. 
Table 5.1 lists the first ten flow values. The first three columns show the 
date and the corresponding riverflow value, $Q$.


# Calculation of the FDC

The flow duration curve is constructed following the calculation steps as seen 
in the right part of the Table 5.1:

- The rank, $i$, of each value (column four in Table 5.1) is calculated (using
the `rank()` function), which means that if the list is sorted, the rank will be its position. Here the series is sorted in descending order and the $i^{th}$ largest value has rank $i$ (i.e. the largest value has rank 1).

- The exceedance frequency (column five in Table 5.1), $EF_{Q_i}$ is calculated as: 
$$EF_{Q_i}  = \frac{i} {N}$$ 
which gives an estimate of the empirical exceedance frequency of the $i^{th}$ largest event. $EF_{Q_i}$ designates here the observed frequency when the flow, $Q$, is larger than the flow value with rank $i$, $Q_i$ .

```{r}
exceedance_frequency <- function(flow)
{
  # current rank
  i <- rank(-flow, ties.method = "min", na.last = "keep") 
  
  # largest rank in sample (= number of non-missing values)
  N <- length(na.omit(flow))
  
  # the exceedance frequency can be seen as the relative rank
  return(i / N)
}

ngaruroro <- ngaruroro %>%
  mutate(
    rank = rank(-discharge, ties.method = "min"), 
    freq.exc = exceedance_frequency(discharge)
  ) 
```

`r tufte::margin_note("Table 5.1 Calculation of daily flow duration curve for River Ngaruroro at Kuripapango, NZ")`
```{r, echo=FALSE}
ngaruroro %>%
  head(7) %>%
  hydroDrought:::export_table(name = "Tab5.1") %>%
  print()
```


```{r, echo=FALSE}
library(kableExtra)
# ngaruroro %>%
#   mutate_if(is.numeric, round, 3) %>%
#   select(Date = time, Flow = discharge, Rank = rank, `Exceedance Frequency` = freq.exc) %>%
#   head(7) %>%
#   kable(caption = "Table 5.1 Calculation of a daily flow duration curve for River Ngaruroro at Kuripapango, NZ") %>%
#   add_header_above(c(" " = 1, "$Q$ in $m^3s^{-1}$" = 1, "$i$" = 1, "$EF_Q$" = 1), 
#                    align = "right")
```


# Plot of the FDC
The sorted table columns are then plotted (Figure 5.2). The ordinate axis is here logarithmic. 

```{r, fig.cap="Figure 5.2 Flow duration curve for River Ngaruroro at Kuripapango, NZ.", fig.margin = FALSE}
ggplot(ngaruroro, aes(x = freq.exc * 100, y = discharge)) +
  geom_line() +
  scale_y_log10(expand = expansion()) +
  scale_x_continuous(expand = expansion()) +
  labs(x = "Exceedance Frequency (%)",
       y = expression(paste("Flow (", m^{3}, s^{-1}, ")"))) + 
  theme(plot.margin = unit(c(0, 10, 0, 0), units = "pt"))
```



# Selected exceedance values

```{r, echo=FALSE}
q90 <- ngaruroro %>%
  filter(freq.exc <= 0.9) %>%
  arrange(desc(freq.exc)) %>%
  head(1) %>%
  pull(discharge)
```

Values for a particular frequency, for example the 90-percentile ($Q_{90}$), can be obtained as the value of $Q$ corresponding to the largest value of $EF_{Q_i}$ that is less than or equal to the value of $EF_{Q_i}$ sought for. A sample of corresponding values in this range is shown in Table 5.2, and the 90-percentile flow value is taken as `r round(q90, 1)`<i>&nbsp;</i>m<sup>3</sup>s<sup>-1</sup>. Alternatively, in case of large differences between successive values, a linear interpolation can be used. 


`r tufte::margin_note("Table 5.2 An extract of values corresponding to $Q_{90}$.")`
```{r echo=FALSE}
tbl <- ngaruroro %>%
  filter(freq.exc >= 0.8999 & freq.exc <= 0.9002) %>%
  arrange(freq.exc) %>%
  hydroDrought:::export_table(name = "Tab5.2") %>%
  print()
```


```{r echo=FALSE}
# tbl  %>%
#   mutate(discharge = round(discharge, 3), 
#          freq.exc = round(freq.exc, 5)) %>%
#   select(Date = time, Flow = discharge, Rank = rank, `Exceedance Frequency` = freq.exc) %>%
#   kable(caption = "Table 5.2 An extract of values corresponding to $Q_{90}$.") %>%
#   add_header_above(c(" " = 1, "$Q$ in ${m^3s^{-1}}$" = 1, "$i$" = 1, "$EF_Q$" = 1), 
#                    align = "right")
```

# Fast-Track

The function `lfquantile()` calculates percentiles (quantiles) directly. The 
exact (interpolated) values for $Q_{95}$, $Q_{90}$ and $Q_{70}$ would be: 

```{r}
lfquantile(ngaruroro$discharge, exc.freq = c(0.95, 0.9, 0.7))
```

The retrieved value for $Q_{90}$ is 
`r lfquantile(ngaruroro$discharge, exc.freq = c(0.9))`<i>&nbsp;</i>m<sup>3</sup>s<sup>-1</sup>, 
approximated to 
`r round(lfquantile(ngaruroro$discharge, exc.freq = c(0.9)), 2)`<i>&nbsp;</i>m<sup>3</sup>s<sup>-1</sup>.

# Standardized Flow Duration Curves

Comparing FDCs from different catchments requires standardization eg. dividing the discharges by the catchment area, the median or the mean discharge. 
```{r}
lambournRay <- international %>%
  filter(river %in% c("Lambourn", "Ray")) %>%
  select(river, area = catchment, data)

lambournRay <- lambournRay %>%
  mutate(
    data = map2(data, area, ~mutate(.x, rel.discharge = discharge * 1000 / .y))
  ) %>%
  print()

fdc <- lambournRay %>%
  select(river, data) %>%
  unnest(cols = data) %>%
  group_by(river) %>%
  mutate(
    freq.exc = exceedance_frequency(rel.discharge)
  ) 
```

```{r, echo=TRUE}
ggplot(fdc, 
       mapping = aes(x = freq.exc * 100, y = rel.discharge, size = river)) +
  geom_step(direction = "vh") +
  scale_x_continuous(expand = expansion()) +
  scale_y_log10(breaks = breaks_log10_all(mult.base = 1),
                minor_breaks = breaks_log10_all(),
                expand = expansion()) +
  scale_size_manual("River", values = c(0.1, 0.75)) +
  labs(x = "Exceedance Frequency (%)",
       y = expression(paste("Flow (", l, s^{-1}, km^{-2}, ")"))) 
```

