---
title: "Mean annual minimum flow"
author: "Worked example 5.2"
---

```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Loading the Data

In this example we are again going to use river flow data from the river Ngaruroro of the `international` dataset in the package **hydroDrought**. 
Ten years of daily data are used as an example, as in Worked Example 5.1, to estimate mean annual minimum of the $n$-day average flow for $n$ equal to 1, 7 and 30 days. For this station the lowest flows are observed around the turn of the calendar year. Therefore the annual minima are selected from **hydrological years** starting 1 September and ending 31 August. Table 5.4 lists the first flow values. 
<!-- The first two columns show the date and the corresponding flow value, Q. -->

In order to calculate the mean annual minimum each observation will be attributed to a hydrological year according to the date of the observation using the function `water_year()` which appends an additional column named `year` to the dataset. 

```{r, warning=FALSE}
library(tidyverse)
library(hydroDrought)

# attribute each observation to the correct hydrological year
# and select only the hydrological years between 1990 und 2000
ngaruroro <- intl %>%
  filter(river == "Ngaruroro") %>%
  select(data) %>%
  unnest(data) %>%
  mutate(
    year = water_year(time, origin = "-09-01")
  ) %>%
  filter(year >= 1990, year <= 1999)

smoothed <- ngaruroro %>%
  mutate(
    MA1 = moving_average(discharge, n = 1),
    MA7 = moving_average(discharge, n = 7),
    MA30 = moving_average(discharge, n = 30)
  )
```

`r tufte::margin_note("Table 5.4 Calculation of n-day average flow (unit: in m<sup>3</sup>s<sup>-1</sup>), River Ngaruroro at Kuripapango, NZ. A moving average with a window length $n$ introduces $n-1$ <code>NA</code> values. ")`
```{r, echo=FALSE}
 smoothed%>%
  print(n = 31)
```


```{r, echo=FALSE}
# num_in_brackets <- function(x) sub("(\\d+)", "(\\1)", x = x)
# library(kableExtra)
# smoothed %>%
#   select(-discharge, -year) %>%
#   head(31) %>%
#   mutate_if(is.numeric, round, digits = 3) %>%
#   rename_all(.funs = num_in_brackets) %>%
#   kable(caption = "Table 5.4 Calculation of n-day average flow (unit: in m<sup>3</sup>s<sup>-1</sup>), River Ngaruroro at Kuripapango, NZ. ") %>%
#   add_header_above(c("Data, 10-year series" = 2, "Moving average calculation" = 2), 
#                    align = "right")
```


```{r}
# compute the annual minima
am <- smoothed %>%
  select(-discharge, -time) %>%
  group_by(year) %>%
  summarise_all(min, na.rm = TRUE)

# average the annual minima to get the mean annual minima
mam <- am %>% 
  select(-year) %>%
  summarise_all(mean) %>%
  print()
```

```{r, echo=FALSE}
# mam  %>%
#   mutate_if(is.numeric, round, digits = 3) %>%
#   rename_all(.funs = num_in_brackets) %>%
#   kable(caption = "Table 5.5 $MAM$(n-day), $n=1$ day, 7 days and 30 days") %>%
#   add_header_above(c("$Q$ in ${m^3s^{-1}}$" = 3), align = "center")
```

# Fast Track

```{r}
ngaruroro %>% 
  select(discharge, time) %>%
  summarise(
    `MAM(1)` = mean_annual_minimum(discharge, time, origin = "-09-01", n = 1),
    `MAM(7)` = mean_annual_minimum(discharge, time, origin = "-09-01", n = 7),
    `MAM(30)` = mean_annual_minimum(discharge, time, origin = "-09-01", n = 30)
  ) %>%
  flatten_dbl()
```

Mapping over the length of the smoothing window avoids the multiple explicit
calls of the function `mean_annual_minimum()`. The results are absolutely 
identical but the code is can be adapted more easily and is less error-prone.


```{r}
c(1, 7, 30) %>%
  map(
    .f = mean_annual_minimum, 
    discharge = ngaruroro$discharge, time = ngaruroro$time, origin = "-09-01"
  ) %>%
  flatten_dbl()
```


