---
title: "Low flow frequency analysis"
author: "Worked example 6.1"
---

```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Loading the Data

River Ngaruroro at Kuripapango in New Zealand (Table 4.3), has been
selected for frequency analysis of annual minimum 1-day values, $AM(1)$,
using the Weibull (WEI) distribution. In mid-latitudes in the Northern
Hemisphere the calendar year can often be used to select the annual
minimum flows. This is a suitable period for the selection of independent
events as the drought or low flow period commonly occurs during the
summer months. In the Southern Hemisphere the low flow season occurs at
the opposite time of the year, and for Ngaruroro the lowest flows are
typically found in the period November to May. As a result $AM(1)$ flows
were selected for a hydrological year starting at the 1<sup>st</sup> September.


```{r}
library(tidyverse)
library(hydroDrought)

ngaruroro <- intl %>%
  filter(river == "Ngaruroro") %>%
  select(data) %>%
  unnest(data) %>%
  # fill gaps with a length of up to 15 days
  sanitize_ts(approx.missing = 15) %>%
   mutate(
    year = water_year(time, origin = "-09-01")
    ) %>%
  print()
```


```{r}
# remove incomplete years
incomplete <- ngaruroro %>%
  filter(!is.na(discharge)) %>%
  pull(time) %>%
  coverage_yearly(origin = "-09-01") %>%
  filter(days.missing > 0) %>%
  print()

ngaruroro <- ngaruroro %>%
    anti_join(incomplete, by = "year")
```


```{r}
# derivation of the annual minima
am <- ngaruroro %>%
  group_by(year) %>%
  summarise(flow = min(discharge)) %>%
  arrange(flow) %>%
  print()
```

```{r, echo=FALSE}
fig.cap <- "Figure 6.9 Time series of annual minimum 1-day flow, AM(1), for River Ngaruroro at
Kuripapango (NZ)."
```
```{r, echo=FALSE, fig.cap=fig.cap}
ggplot(am, aes(x = year, y = flow)) + 
  geom_point() + 
  geom_smooth(method = "loess", formula = "y ~ x", span = 1, se = FALSE,
              col = "black", size = 0.2, linetype = "dashed") + 
  labs(y = expression(paste("Annual minimum flow (", m^{3}, s^{-1}, ")"))) +
  scale_y_continuous(limits = c(0, NA)) + 
  theme(axis.title.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

```{r, echo=FALSE, }
fig.cap <- "Sample annual minimum 1-day flow, AM(1), for River Ngaruroro at
Kuripapango (NZ); histogram"
```
```{r, echo=FALSE, fig.cap=fig.cap}
ggplot(am, aes(flow)) + 
  geom_histogram(bins = 8, boundary = 0, binwidth = 0.5, col = "white") +
  labs(x = expression(paste("Flow classes (", m^{3}, s^{-1}, ")")),
       y = "Number of observations in each class") 
```


Ngaruroro has a flashy river regime, but has no observed zero flows. The
$AM(1)$ flows are considered to come from the same population as only a
minor part of the catchment is influenced by snow in winter. The
observations cover the period `r min(am$year)` to `r min(am$year)` and with the omission of 
`r nrow(incomplete)` 
years (`r hydroDrought:::paste_with_and(incomplete$year)`) with missing data a total of `r nrow(am)` values results. A histogram of the
values is shown in Figure 6.3 (upper left). To test the assumption of
stationarity the $AM(1)$ values are plotted against time in Figure 6.9. No
trend can be detected in the series and the data are therefore assumed to
fulfil the requirement of independent and identically distributed data (iid).


# Derivation of an empirical distribution function

```{r}
empirical <- am %>%
  mutate(
    rank = rank(flow),
    prob.emp = rank / (n() + 1),
    rp.emp = 1 / prob.emp
  ) %>%
  arrange(rank)  %>%
  print()
```

a) The $x$ values, AM(1), are sorted in ascending order and the rank of
each value is calculated. The smallest value equals `r min(am$flow)` in m<sup>3</sup>s<sup>-1</sup> and is given rank 1.

b) The non-exceedance probability, $F(x) = p$, (column `prob.emp`) is calculated for each $x$
value using the Weibull plotting position formula $p_i = i / (n + 1)$ (Equation 6.6).

c) A probability plot for the $AM(1)$ values is obtained by plotting the
flow values against $F(x)$ as demonstrated in Figure 6.10 (upper). A
staircase pattern of several nearly equal values is observed. (Note that
the $AM(1)$ values now are plotted on the $y$-axis as compared to
Figure 6.3, lower).

```{r, echo = FALSE}
# ggplot(empirical, aes(x = prob.emp, y = flow)) + 
#   geom_point()
```

d) The return period (column `rp.emp`) of the smallest event can be calculated following
Equation 6.5. The non-exceedance frequency, $F(x)$, of the smallest
event equals `r round(empirical$prob.emp[1], 3)` according to step 2(b), which gives a return period
of `r empirical$rp.emp[1]`  years.


# Fitting a distribution  
Fitting the two-parameter Weibull (WEI) distribution function
(Section A6.1.4) using the method of L-moments and the fact that
$Y = −\ln(X)$ has a Gumbel distribution if $X$ has a WEI distribution. 
^[The direct estimation of the Weibull parameters is shown in [section Fast Track](#ft). ]


a) The first two L-moments ($\hat\lambda_1$ and $\hat\lambda_2$) are estimated based on time series
of $\ln(x)$. These so called sample L-moments are calculated with the function `samlmu()` from the package **lmom**.

```{r}
library(lmom)
moments <- samlmu(log(am$flow))
moments 
```

b) The parameter estimates of the WEI distribution (shape parameter $\kappa$, scale parameter $\alpha$) are obtained
following Stedinger et al. (1993):

$\hat\kappa = \frac{\ln(2)}{\hat\lambda_2}$ and $\hat\alpha = \exp\left( \hat\lambda_1 + \frac{0.5772}{\hat\kappa}\right)$ 

```{r, echo=FALSE}
# values of the first edition
# kappa <- 5.776
# alpha <- 4.441

# we can reproduce the numbers from the first edition but they differ from the 
# results by lmom-package. Especially the shape parameter kappa is off
# with kappa <- 5.502 we would get approximately the same results as 
```

^[Tobias: The shape parameter $\hat\kappa$ differns quite a lot from the result by the lmom-package ($\hat\kappa = 5.502302$). I cannot figure out why thats the case. Susbsequently the quantile and probability estimates in section Fast-Track are also slightly off.]


```{r}
kappa <- log(2) / unname(moments["l_2"])
alpha <- exp(unname(moments["l_1"]) + 0.5772 / kappa)
parameter <- c(location = 0, scale = alpha, shape = kappa)
parameter
```

c) $F(x)$ for the WEI distribution is obtained following Equation A6.1.33
(recalling that location parameter $\xi$ is set to zero):

$$ F(x) = 1 - \exp\left[-\left(\frac{x}{\alpha}\right)^\kappa\right]$$
```{r}
fitted <- empirical %>%
  select(year, flow, prob.emp) %>%
  mutate(
    prob.wei = 1 - exp(-(flow / alpha)^kappa),
    rp.wei = 1 / prob.wei
  ) 
```


```{r, echo=FALSE}
fig.cap <- "Figure 6.10 (upper): Estimated quantiles for annual minimum 1-day flow, AM(1), for River
Ngaruroro at Kuripapango (NZ); probability plot showing the Weibull distribution
(curve)."
```
```{r, echo=FALSE, fig.cap=fig.cap}
# quant.theo = qweibull(empirical, shape = kappa, scale = alpha),
# quant.theo2 = exp(log(-log(1-empirical))/kappa + log(alpha))

# tocheck: we could estimate the fitted distribution much smoother... 
ggplot(fitted, aes(prob.emp, flow)) + 
  geom_point() + 
  geom_line(aes(x = prob.wei)) + 
  scale_y_continuous(limits = c(0, NA)) + 
  labs(x = expression(paste("F(", x, ")")),
       y = expression(paste("Annual minimum flow (", m^{3}, s^{-1}, ")")))


```


```{r, echo=FALSE}
fig.cap <- "Figure 6.10 (lower): QQ-plot of empirical quantiles versus Weibull quantiles"
```
```{r, echo=FALSE, fig.cap=fig.cap}
# todo: in the old book, these are not quantiles! 
ggplot(fitted, aes(x = prob.emp, y = prob.wei)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0, size = 0.2, linetype = "dashed") + 
  coord_fixed()
```
^[Tobias: The plot doesn't show quantiles (as mentioned in the first edition where it is called a QQ-Plot), but probabilities... ]

d) The data can be plotted in a probability plot and compared to the
empirical quantiles in step 2(b) (Figure 6.10, upper). The plot shows
that the WEI distribution is well adjusted to the low flow extreme
values, whereas the deviation in the upper range might suggest that the
three highest values do not belong to the low flow population
(Section 6.3.3). In Figure 6.10 (lower) the empirical quantiles are
plotted against the estimated distribution quantiles (qq-plot). The
points should be close to the unit diagonal if the data fit the WEI
distribution well. The use of a plotting position implies that the
ordered sample is plotted in regularly spaced positions. As
demonstrated in the figure, the observed jumps in $AM(1)$ values are
reflected in the estimated Weibull quantiles, but not in the empirical
quantiles.

e) The non-exceedance frequency of the smallest value equals 
`r round(fitted$prob.wei[1], 4)` (Equation A6.1.33), which gives a return period of 
`r round(fitted$rp.wei[1], 1)` years (Equation 6.5). 
This is `r round(empirical$rp.emp[1] - fitted$rp.wei[1], 1)` 
years less that the empirical estimate derived in step 2(d).

```{r}
fitted
```


f) The 50 and 100-year events can be estimated from Equation A6.1.34:
$$ x_p = \xi + \alpha \left[-\ln( 1 - p ) \right]^{1/\kappa} $$

```{r}
quantile_weibull <- function(prob, location = 0, scale, shape) {
  location + scale * (-log(1 - prob))^(1/shape)
}

return.period <- c(50, 100)
q <- quantile_weibull(prob = 1/return.period, scale = alpha, shape = kappa) 

round(q, 2)
```

which gives: $\hat x_{50} = `r format(q[1], digits = 2, nsmall = 2)`$m<sup>3</sup>s<sup>-1</sup> 
and $\hat x_{100} = `r format(q[2], digits = 2, nsmall = 2)`$m<sup>3</sup>s<sup>-1</sup>. 


# Fast Track {#ft}

To obtain L-moment estimates of the two parameter Weibull distribution the 
lower bound (location parameter $\xi$) is set to zero otherwise a three parameter
distribution is fitted.
```{r}
library(lmom)
lmom <- samlmu(am$flow)
par.wei <- pelwei(lmom, bound = 0)
par.wei
```

Quantiles for given probabilities can be calculated for many suitable 
distribution by the corresponding `qua...()` functions. 

```{r}
quawei(f = c(0.02, 0.01), para = par.wei) %>%
  round(digits = 2)
```

# Probability plot using the exponential reduced variate
```{r}
fitted.fast <- am %>%
  mutate(
    prob.emp = rank(flow) / (n() + 1),
    prob.wei = cdfwei(x = flow, para = pelwei(lmom, bound = 0)),
    prob.exp = cdfexp(x = flow, para = pelexp(lmom))
  ) %>% 
  arrange(flow) %>%
  print()
```

a) For comparison of the two-parameter EXP (Equation A6.1.64) and
WEI distribution the exponential reduced variable is applied
(Box 6.4). By substituting $y$ into the expression for $F(x)$, $y$ can be
expressed as $-\ln(1 - F(x))$ (Equation B6.4.6). The parameters of the
EXP distribution are estimated using L-moments (Equation A6.1.72)
and used to calculate $y$ and subsequently $F(x)$.

b) The non-exceedance probability $F(x)$ for the observations is
determined using the Weibull plotting position formula
(Equation 6.6).

c) The $AM(1)$ values are plotted against the reduced variate in
Figure 6.11. A reduced variate of $0.02$ corresponds to a return period
of $50.5$ years for minimum values (Equation 6.5). The data will plot as
a straight line given they follow the EXP distribution. Again it is
demonstrated that the WEI distribution fits the extreme low flow
values well, and also the upper range apart from the three largest
values. The two-parameter EXP distribution is less suited to model the
sample. Alternatively, $\ln(X)$ could be plotted on a Gumbel
probability paper, and a straight line would result provided the data
fitted the WEI distribution.

```{r, echo=FALSE}
fig.cap <- "Figure 6.11: The annual minimum 1-day flow, AM(1), plotted against the reduced variate, y, of the EXP distribution for River Ngaruroro at Kuripapango (NZ); the observations marked as points, the two-parameter EXP distribution (continous line) and the two-parameter WEI distribution (dashed line)."
```


```{r, echo=FALSE, warning=FALSE, fig.cap=fig.cap}
reduced_var <- function (x) -log(1 - x)

d <- fitted.fast %>%
  select(year, flow, prob.wei, prob.exp) %>%
  pivot_longer(c(prob.wei, prob.exp), 
               names_to = "Distribution", values_to = "probability") %>%
  mutate(Distribution = c(prob.wei = "Weibull", prob.exp = "Exponential")[Distribution])

ggplot(empirical, aes(x = reduced_var(prob.emp), flow)) + 
  geom_point() + 
  geom_line(data = d, aes(x = reduced_var(probability), 
                          col = Distribution, linetype = Distribution)) +
  scale_x_continuous(limits = c(0.01, 4)) +
  labs(x = expression(paste("Exponential reduced variate -log(", 1, " - F(", x, "))")),
       y = expression(paste("Annual minimum flow (", m^{3}, s^{-1}, ")")))
```


