---
title: "Linear trend analysis"
author: "Worked example 7.1"
---

```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Loading the Data

We begin by loading the Elbe and Gota flow time series and then calculating annual minimum flow for each. We are purposefully choosing to use the calendar year (beginning January 1) for this analysis rather than water year (beginning October 1 or November 1 for many northern hemisphere countries) because the water year breakpoint occurs near the seasonal minimum flow.

```{r}
library(tidyverse)
library(hydroDrought)
library(lubridate)

### Filter to the rivers and create columns for dates
flow_df <- international %>%
	filter(river == "Elbe" | river == "Gota") %>%
	select(river, data) %>%
	unnest(cols = data) %>%
	mutate(
		year = year(time),
		jdate = yday(time)
	)  %>%
	print()

### Group by year and river and then calculate min and mean flow
flow_annual <- flow_df %>%
	group_by(river, year) %>%
	summarise(min_m3s = min(discharge, na.rm=TRUE), mean_m3s=mean(discharge, na.rm=TRUE), .groups = "drop_last") %>%
	print()
```

```{r, echo=FALSE}
plot_df <- flow_df %>% filter((river == "Gota" & year < 1935) | river == "Elbe")

### Double check plot of seasonal pattern 
ggplot(plot_df, aes(x=jdate, y=discharge)) %>%
	+ geom_point(alpha =0.2) %>%
	+ geom_smooth(method = "gam", formula = y ~ s(x, bs = "cc")) %>%
	+ geom_vline(xintercept=yday(c("2000-01-01", "2000-09-01", "2000-10-01")), colour = "red") %>%
	+ facet_grid(river ~ . , scales = "free_y")
```
# Linear trend for Elbe River

Consider the annual minimum flow for the Elbe river (Fig. 7.8). By fitting a line of form Eq 7.43, the estimated slope is -0.0408 m3/s per year, with a 95% confidence interval of -0.34 to 0.26 m3/s per year. As you can see, we are unclear whether the true trend is positive or negative and this is reflected in a p-value of 0.79 for the t-test of the slope coefficients. Because this p-value is well above 0.05, we accept the null hypothesis (Ho) that there is likely no significant trend in annual minimum flow for the Elbe River. 

```{r}
### Filter to only the Elbe River
elbe_annual <- flow_annual %>%
	filter(river == "Elbe") 

### Create an OLS linear trend model 
elbe_lm <- lm(min_m3s~year, data = elbe_annual)

### Show results
summary(elbe_lm)
```

```{r}
### Insert columns for trend fit and residuals
elbe_annual <- elbe_annual %>%
	mutate(fitted = fitted(elbe_lm)) %>%
	mutate(resid = resid(elbe_lm)) %>%
	print()

### Create dataframe for confidence interval and predition interval
new_data <- data.frame(year = seq(1865, 2025, 1))
elbe_conf <- predict(elbe_lm, newdata=new_data, interval = "confidence", level = 0.95) %>%
	as.data.frame() %>%
	bind_cols(new_data) %>%
	mutate(interval = "Confidence", level = 0.95) %>%
	print()

elbe_pred <- predict(elbe_lm, newdata=new_data, interval = "prediction", level = 0.95) %>%
	as.data.frame() %>%
	bind_cols(new_data) %>%
	mutate(interval = "Prediction", level = 0.95)
```


```{r, echo=FALSE}
fig.cap <- "Figure 7.8: Linear trend of annual minimum flow on the Elbe River at Neu Darchau. Light region is the prediction interval, while the darker region is the confidence interval, both using a 95% interval."

plot_ribbon <- elbe_conf %>%
	bind_rows(elbe_pred)

plot_line <- new_data %>%
	mutate(line_fit = predict(elbe_lm, newdata=new_data), Data = "Linear") 
```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot annual min flows and the linear regression
p <- ggplot(elbe_annual, aes(x=year)) %>%
	+ geom_ribbon(data = plot_ribbon, aes(ymin= lwr, ymax = upr, fill = interval), alpha = 0.5) %>%
	+ geom_point(aes(y=min_m3s), alpha = 0.5, colour = "black") %>%
	+ geom_line(data = plot_line, aes(y=line_fit, colour = Data), alpha = 0.5) %>%
	+ theme_classic() %>%
	+ scale_colour_manual(name = "", values = c("blue")) %>%
	+ scale_fill_manual(name = "Interval (95%)", values = c("grey25", "grey83")) %>%
	+ scale_x_continuous( name="Year", breaks=seq(0,3000,10)) %>%   ## Set the x-axis title and style to be years with breaks every 10 years
	+ scale_y_continuous(name = bquote('Annual Min Flow'~(m^3/s)), labels = scales::comma) %>% ## Set the y-axis title and add commas to values
	+ coord_cartesian(xlim=c(1869,2021), ylim = c(90,550), expand = FALSE) %>%
	+ theme(legend.position=c(0.93,0.9), legend.spacing = unit(0.15, "cm")) %>%
	+ guides(colour = guide_legend(title.position="top", title.hjust = 0.1, order = 1),
         fill = guide_legend(title.position="top", title.hjust = 0.1, order = 2))	

### Display figure (stored in variable p)
p
```

# Linear trend for Göta River

If we instead consider the annual minimum flow on the Göta River, the annual trend is 1.23 m3/s per year (Fig. 7.9 with a 95% confidence interval of 0.875 to 1.59 m3/s per year. The t-test on this coefficient produces a p-value of 4 x 10-9, meaning it is highly unlikely that such a strong trend would have appeared randomly in observations if there were no true underlying trend. We reject the null hypothesis, and accept the alternative hypothesis, that there is a significant trend in Göta River annual minimum flows. Remember that we set up our alternative hypothesis as a two-sided test (b1 ≠ 0), so we are testing whether there is a significant trend, regardless if it is positive or negative. We should make this clear when we report our findings. Of course, we should also report that the trend we found was that Göta River minimum flows have been increasing at an estimated rate of 1.23 m3/s per year since 1941. 

```{r}
### Filter to the Gota River and years after 1940 
gota_annual <- flow_annual %>%
	filter(river == "Gota") %>%
	filter(year >= 1941)

### Create an OLS linear trend model 
gota_lm <- lm(min_m3s~year, data = gota_annual)

### Show results
summary(gota_lm)
```



```{r, echo=FALSE}
### Insert columns for trend fit and residuals
gota_annual <- gota_annual %>%
	mutate(fitted = fitted(gota_lm)) %>%
	mutate(resid = resid(gota_lm)) %>%
	print()

### Create dataframe for confidence interval and predition interval
new_data <- data.frame(year = seq(1935, 2025, 1))
gota_conf <- predict(gota_lm, newdata=new_data, interval = "confidence", level = 0.95) %>%
	as.data.frame() %>%
	bind_cols(new_data) %>%
	mutate(interval = "Confidence", level = 0.95) %>%
	print()

gota_pred <- predict(gota_lm, newdata=new_data, interval = "prediction", level = 0.95) %>%
	as.data.frame() %>%
	bind_cols(new_data) %>%
	mutate(interval = "Prediction", level = 0.95)

fig.cap <- "Figure 7.9: Linear trend of annual minimum flow on the Gota River. Light region is the prediction interval, while the darker region is the confidence interval, both using a 95% interval."

plot_ribbon <- gota_conf %>%
	bind_rows(gota_pred)

plot_line <- new_data %>%
	mutate(line_fit = predict(gota_lm, newdata=new_data), Data = "Linear") 
```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot annual min flows and the linear regression
p <- ggplot(gota_annual, aes(x=year)) %>%
	+ geom_ribbon(data = plot_ribbon, aes(ymin= lwr, ymax = upr, fill = interval), alpha = 0.5) %>%
	+ geom_point(aes(y=min_m3s), alpha = 0.5, colour = "black") %>%
	+ geom_line(data = plot_line, aes(y=line_fit, colour = Data), alpha = 0.5) %>%
	+ theme_classic() %>%
	+ scale_colour_manual(name = "", values = c("blue")) %>%
	+ scale_fill_manual(name = "Interval (95%)", values = c("grey25", "grey83")) %>%
	+ scale_x_continuous( name="Year", breaks=seq(0,3000,10)) %>%   ## Set the x-axis title and style to be years with breaks every 10 years
	+ scale_y_continuous(name = bquote('Annual Min Flow'~(m^3/s)), labels = scales::comma) %>% ## Set the y-axis title and add commas to values
	+ coord_cartesian(xlim=c(1940,2021), ylim = c(-10,310), expand = FALSE) %>%
	+ theme(legend.position=c(0.1,0.9), legend.spacing = unit(0.15, "cm")) %>%
	+ guides(colour = guide_legend(title.position="top", title.hjust = 0.1, order = 1),
         fill = guide_legend(title.position="top", title.hjust = 0.1, order = 2))	

### Display figure (stored in variable p)
p
```

```{r, echo=FALSE}
### Plot the residuals
p <- ggplot(gota_annual, aes(x=year, y=resid)) %>%
	+ geom_point(colour="grey50") %>%  ## Plot as points. 
	+ stat_summary(fun.data=mean_cl_normal) %>%
	+ geom_hline(yintercept=0, linetype="dashed", color = "grey50") %>%
	+ theme_classic() %>%  ## Use the classic theme (vertical and horizontal axes)
	+ scale_x_continuous(name="Year", breaks=seq(0,3000,10)) %>%   ## Set the x-axis title and style to be years with breaks every 10 years
	+ scale_y_continuous(name=bquote('OLS Trend Residual'~(m^3/s)), labels = scales::comma) ## Set the y-axis title and add commas to values
p
```


