---
title: "Non-Parametric Trends"
author: "Worked example 7.2"
---

```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Loading the Data
First we must load the data. Consider the same data as Worked Example 7.1 using annual minimum flow from the Göta River beginning in 1941.

```{r}
library(tidyverse)
library(hydroDrought)
library(lubridate)
#require(Hmisc)
require(zyp)

### Filter to the rivers and create columns for dates
flow_df <- international %>%
	filter(river == "Gota") %>%
	select(river, data) %>%
	unnest(cols = data) %>%
	mutate(
		year = year(time),
		jdate = yday(time)
	)  %>%
	print()

### Group by year and river and then calculate min annual flow
flow_annual <- flow_df %>%
	group_by(river, year) %>%
	summarise(min_m3s = min(discharge, na.rm=TRUE), .groups = "drop_last") %>%
	print() %>%
	ungroup()
```

```{r}
### Subset to 1941 and after
flow_annual_subset <- flow_annual %>% filter( year >= 1941) 

### Run the Mann-Kendall Test for the annual min flow
mk_test_results <- Kendall(x=flow_annual_subset$year, y=flow_annual_subset$min_m3s)
summary(mk_test_results)
```
# Mann-Kendall test

The Mann-Kendall test produces an S value of `r as.numeric(mk_test_results$S)`derived by calculating the difference between 2200 positive pairs (73.3%) and 774 negative pairs (25.8%) out of 3003 possible data pairs (29 pairs had equal values). Using this test statistic and the `r dim(flow_annual_subset)[1]` years of observations, the Mann-Kendall test produces a p-value < 1x10-15 suggesting there was a statistically significant increase in annual minimum flow during this period. This finding agrees with our linear regression trend test, though with slightly less confidence because we gave up some statistical power for the flexibility of a non-parametric test.

As discussed in the text, the Mann-Kendall test only considers increases/decreases and does not estimate the magnitude of the underlying trend. For this, we must calculate the Theil-Sen non-parametric line. When compared to the OLS estimate (Fig. 7.10) the Theil-Sen line shows very similar results, but is slightly less affected by relatively high outliers during the late 2000s. 

```{r}
### Run the Theil-Sen slope for annual min flow
sen_slope_results <- zyp.sen(min_m3s ~ year, data = flow_annual_subset)
sen_slope_results
confint.zyp(sen_slope_results)
```

```{r}
### Rerun OLS trend
gota_lm <- lm(min_m3s~year, data = flow_annual_subset)

### Combine for plotting purposes
trend_lines <- flow_annual_subset %>% 
	select(year) %>%  ### Only select the year
	mutate(   ### Create two new columns with predictions
		ols = gota_lm$coef[1] + gota_lm$coef[2] * year,
		sen = sen_slope_results$coef[1] + sen_slope_results$coef[2] * year
	) %>% 
	gather(model, min_m3s, -year)   ### Gather these predictions into a long format for plotting

### Re-factor the model names to make it better to plot
trend_lines$model <- factor(trend_lines$model, c("ols", "sen"), labels=c("Ordinary Least Squares (OLS)", "Thiel-Sen Slope"))
```

```{r, echo=FALSE}
fig.cap <- "Figure 7.10: Comparison of non-parametric (Theil-Sen) trend with parametric (OLS) trend for annual minimum flows on the Gota river after 1940."
```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot annual min flows and the linear regression
p <- ggplot(flow_annual_subset, aes(x=year, y=min_m3s)) %>%
	+ geom_point(aes(y=min_m3s), alpha = 0.5, colour = "black") %>%
	+ geom_line(data = trend_lines, aes(colour=model, linetype=model)) %>%  ### Plot line fits
	+ theme_classic() %>%  ## Use the classic theme (vertical and horizontal axes)
	+ scale_colour_brewer(name="Model", type="qual", palette="Set2")  %>%  ### Set colors, use color-deficient friendly palette
	+ scale_linetype_manual(name="Model", values=c("longdash","solid"))  %>%   ### Set linetypes
	+ scale_x_continuous(name="Year", breaks=seq(0,3000,10)) %>%   ## Set the x-axis title and style to be years with breaks every 10 years
	+ scale_y_continuous(name = bquote('Annual Min Flow'~(m^3/s)), labels = scales::comma) %>% ## Set the y-axis title and add commas to values
	+ theme(legend.position = c(0.25, 0.9))   ## Move legend to inside figure, upper left
p

```
# High leverage outlier effects

If we instead expand the Göta River timeseries to begin in 1934, we introduce three high leverage outliers during the late 1930s. Here, the robustness of the Theil-Sen line is more apparent. 

```{r}
### Subset to 1934 and after
flow_annual_longer <- flow_annual %>% filter( year >= 1934) 

### Run the Mann-Kendall Test for the annual min flow
mk_longer <- Kendall(x=flow_annual_longer$year, y=flow_annual_longer$min_m3s)
summary(mk_longer)

### Run the Theil-Sen slope for annual min flow
sen_slope_longer <- zyp.sen(min_m3s ~ year, data = flow_annual_longer)
sen_slope_longer
confint.zyp(sen_slope_longer)

### Rerun OLS trend
gota_lm_longer <- lm(min_m3s~year, data = flow_annual_longer)
summary(gota_lm_longer)
```
By including only 6 additional years, the OLS line changes from a significantly positive trend (+`r gota_lm$coef[2]` m3/s per year) to a significantly negative trend (`r gota_lm_longer$coef[2]` m3/s per year). The Theil-Sen line is slightly affected by these outliers, decreasing in slope from `r sen_slope_results$coef[2]` to `r sen_slope_longer$coef[2]`, but maintaining a statistically significant 70-year increasing trend that is clear visually. 


This is an extreme example and we know from Section 7.2.3 that these outliers are related to a change in lake management. Data before 1940 follows a different underlying distribution, violating the assumptions of our model. To prove this, read ahead to Section 7.4.3. 

```{r}
### Combine for plotting purposes
trend_lines <- flow_annual_longer %>% 
	select(year) %>%  ### Only select the year
	mutate(   ### Create two new columns with predictions
		ols = gota_lm_longer$coef[1] + gota_lm_longer$coef[2] * year,
		sen = sen_slope_longer$coef[1] + sen_slope_longer$coef[2] * year
	) %>% 
	gather(model, min_m3s, -year)   ### Gather these predictions into a long format for plotting

### Re-factor the model names to make it better to plot
trend_lines$model <- factor(trend_lines$model, c("ols", "sen"), labels=c("Ordinary Least Squares (OLS)", "Thiel-Sen Slope"))
```

```{r, echo=FALSE}
fig.cap <- "Figure 7.11: Comparison of non-parametric (Theil-Sen) trend with parametric (OLS) trend for annual minimum flows on the Gota river by incorrectly including data from 1934 (prior to dam regulation)."
```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot annual min flows and the linear regression
p <- ggplot(flow_annual_subset, aes(x=year, y=min_m3s)) %>%
	+ geom_point(aes(y=min_m3s), alpha = 0.5, colour = "black") %>%
	+ geom_line(data = trend_lines, aes(colour=model, linetype=model)) %>%  ### Plot line fits
	+ theme_classic() %>%  ## Use the classic theme (vertical and horizontal axes)
	+ scale_colour_brewer(name="Model", type="qual", palette="Set2")  %>%  ### Set colors, use color-deficient friendly palette
	+ scale_linetype_manual(name="Model", values=c("longdash","solid"))  %>%   ### Set linetypes
	+ scale_x_continuous(name="Year", breaks=seq(0,3000,10)) %>%   ## Set the x-axis title and style to be years with breaks every 10 years
	+ scale_y_continuous(name = bquote('Annual Min Flow'~(m^3/s)), labels = scales::comma) %>% ## Set the y-axis title and add commas to values
	+ theme(legend.position = c(0.25, 0.9))   ## Move legend to inside figure, upper left
p
```

