---
title: "Structural Change Analysis"
author: "Worked example 7.3"
---

```{r, setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Visualizing Daily Flow
In previous worked examples, we used a subset of Göta River discharges, processed to produce the annual minimum flow cut either to the period before 1940 or after 1940. Plotting the raw daily data from the earliest records (Fig. 7.12), we see an abrupt and obvious structural change around 1940. There is clearly a change in variance and a change in temporal autocorrelation (flow persistence) at this point in time. 

```{r}
library(tidyverse)
library(hydroDrought)
library(lubridate)
library(strucchange)
library(Hmisc)

### Filter to the rivers and create columns for dates
flow_df <- international %>%
	filter(river == "Gota") %>%
	select(river, data) %>%
	unnest(cols = data) %>%
	mutate(
		year = year(time),
		jdate = yday(time)
	)  %>%
	print()
```

```{r, echo=FALSE}
fig.cap <- "Figure 7.12: Daily discharge at the Gota River station demonstrating a visually clear structural change."
```

```{r, echo=FALSE, fig.cap=fig.cap}
### Plot annual min flows and the linear regression
date_breaks <- seq(from = as.Date("1850-01-01"), to = as.Date("2050-01-01"), by = "10 years")

p <- ggplot(flow_df, aes(x=time, y=discharge)) %>%
	+ geom_line(colour = "grey20") %>%
	+ theme_classic() %>%  ## Use the classic theme (vertical and horizontal axes)
	+ scale_x_date(name = "Year", breaks = date_breaks, date_labels = "%Y") %>%
	#+ scale_x_continuous(name="Date", breaks=seq(0,3000,10)) %>%   ## Set the x-axis title and style to be years with breaks every 10 years
	+ scale_y_continuous(name = bquote('Daily Discharge'~(m^3/s)), labels = scales::comma) 
p
```

# Testing for a change point

To avoid this shift in temporal autocorrelation, let us focus on the annual minimum flow for the entire time series. We see a severe decrease in flow from 300-600 m3 s-1prior to the 1930s and 50-200 m3 s-1 after 1940 (Fig. 7.13). Note, that examples are not normally this clear. We can apply the Chow F-test to identify the change point. 

```{r}
### Group by year and river and then calculate min annual flow
flow_annual <- flow_df %>%
	group_by(river, year) %>%
	summarise(min_m3s = min(discharge, na.rm=TRUE), .groups = "drop_last") %>%
	print() %>%
	ungroup()
```

```{r, echo=FALSE}
fig.cap <- "Figure 7.13: Annual minimum discharge at the Gota River station for the full period (1850-present)."
```

```{r, echo=FALSE, fig.cap=fig.cap}
p <- ggplot(flow_annual, aes(x=year, y=min_m3s)) %>%
	+ geom_line(colour = "grey20", alpha=0.5) %>%
	+ geom_point(alpha = 0.5, colour = "black") %>%
	+ theme_classic() %>%  ## Use the classic theme (vertical and horizontal axes)
	+ scale_x_continuous(name="Year", breaks=seq(0,3000,10)) %>%   ## Set the x-axis title and style to be years with breaks every 10 years
	+ scale_y_continuous(name = bquote('Annual Minimum Discharge'~(m^3/s)), labels = scales::comma) 
p
```

```{r}
### Convert the data to a regularly spaced annual time series
first_year <- min(flow_annual$year, na.rm=TRUE)

flow_annual_ts <- flow_annual %>% 
	complete(year=full_seq(year,1)) %>%    ### Make sure all years are included using the complete command
	ts(frequency=1, start=first_year) 

### Run the Chow F test for a single breakpoint, assuming there is no trend in the data (ie intercept only)
### The equation therefore looks like min_m3s is a function only of an intercept (designated by a 1)
f_test_int_only <- Fstats(min_m3s ~ 1, data=flow_annual_ts)
sctest(f_test_int_only)
bp_int_only <- breakpoints(f_test_int_only)
summary(bp_int_only)
```

```{r, include =FALSE}
plot(f_test_int_only, alpha=0.05)
lines(bp_int_only) 
```

Fig. 7.14 shows the null model in green, assuming the data come from the same distribution, and the alternative model in blue, assuming a sudden change point. In this example, we choose to constrain the model to assume there is a constant mean with no trends, i.e. b1=b2a=b2b=0. Using a single breakpoint, we see that the F-test statistic is maximized with a breakpoint in 1937. This breakpoint is highly statistically significant, suggesting there was a detectable structural change during this year. The 95% confidence interval around the location of this breakpoint is between 1936 and 1938, also giving us strong confidence in the location of this breakpoint. 


```{r}
### Run for multiple potential breakpoints
all_bp_int_only <- breakpoints(min_m3s ~ 1, data=flow_annual_ts)
summary(all_bp_int_only)

### Best solution should minimize BIC (Bayesian Information Criterion)
#plot(all_bp_int_only)

### Calculate a confidence interval around this
ci_int_only <- confint(all_bp_int_only)
ci_int_only

### Assume this breakpoint and plot the null hypothesis model (no breakpoint, intercept only) vs the alternative (breakpoint, intercept only)
### Fit both models
fm0 <- lm(min_m3s ~ 1, data=flow_annual_ts )
fm1 <- lm(min_m3s ~ breakfactor(all_bp_int_only, breaks = 1) + 1, data=flow_annual_ts )
```

```{r, echo=FALSE}
fig.cap <- "Figure 7.14: Candidate models for Chow Breakpoint Test. Null model (no breakpoint) shown in green, while the alternative model (single breakpoint) is shown in blue. Location of the breakpoint shown with a dashed line and a 95% confidence interval in red."
```

```{r, echo=FALSE, fig.cap=fig.cap}
plot(flow_annual_ts[,"min_m3s"], xlab="Date", ylab="Annual Minimum Flow (m3s)")
lines(ts(fitted(fm0), start = first_year), col = 3)
lines(ts(fitted(fm1), start = first_year), col = 4)
### Add a line for breakpoint
lines(bp_int_only)
### Add confidence interval around breakpoint
lines(ci_int_only)
```

To test the possibility of additional breakpoints, we introduce more complex models with 2-5 breakpoints (Bai and Perron, 2003; Zeileis et al. 2003). However, adding these breaks did not significantly decrease the model residuals, pointing to the presence of a single breakpoint.

As with all statistical findings, it is important to validate results with real-world understanding, checking whether the findings make physical sense. In this case, we can check the historical record during the period 1936–38 to try and identify what physical processes could produce this sudden decrease in annual minimum flows on the Göta River. The Lake Vänern decree was signed in 1937, which codified water management on Lake Vanern, the headwaters of Göta River, to facilitate hydropower. This more active reservoir management focused on maintaining a regular pattern of seasonal flow characterized by low summer flows (150 m3 s-1) and high winter flows (900 m3 s-1), with occasional fluctuation between these two extremes. Prior to management, lake flows tended to increase and decrease along decadal time scales with narrower variance. In this way, the structural change both decreased annual low flows and increased peak flows to maximize hydropower. 


```{r, include=FALSE}
### Run the Chow F test for a single breakpoint, allowing for a trend
f_test_slope_int <- Fstats(min_m3s ~ year + 1, data=flow_annual_ts)
sctest(f_test_slope_int)
bp_slope_int <- breakpoints(f_test_slope_int)
summary(bp_slope_int)
### F statistics indicate a significant breakpoint. The maximum F statistic (most severe breakpoint) occurs at 1956

### Plot the value of the F statistic, add in a red line for alpha = 0.05, rejection criteria
plot(f_test_slope_int, alpha=0.05)
lines(bp_slope_int)  ### Add line for the maximum F-Value
### We can use this plot to visually see if there are more than 1 potential break points

### Run for multiple potential breakpoints
all_bp_slope_int <- breakpoints(min_m3s ~ year +1, data=flow_annual_ts)

### Best solution should minimize BIC (Bayesian Information Criterion)
plot(all_bp_slope_int)
summary(all_bp_slope_int)
### Still looks like a single breakpoint is the best (minimizes BIC)

### Calculate a confidence interval around this
ci_slope_int <- confint(all_bp_slope_int)
ci_slope_int
### Wider confidence interval

### Assume this breakpoint and compare the null hypothesis model (no breakpoint, slope intercept model) vs the alternative (breakpoint, slope intercept)
### Fit both models
fm0 <- lm(min_m3s ~ year + 1, data=flow_annual_ts )
fm2 <- lm(min_m3s ~ breakfactor(bp_slope_int, breaks = 1)*year + 1, data=flow_annual_ts )

### Plot the results
plot(flow_annual_ts[,"min_m3s"])
lines(ts(fitted(fm0), start = first_year), col = 3)
lines(ts(fitted(fm2), start = first_year), col = 4)
### Add a line for breakpoint
lines(bp_slope_int)
### Add confidence interval around breakpoint
lines(ci_slope_int)
	 
### For comparison, include the previous breakpoint model without a slope
fm1 <- lm(min_m3s ~ breakfactor(bp_int_only, breaks = 1) + 1, data=flow_annual_ts )

### Plot the results
plot(flow_annual_ts[,"min_m3s"])
lines(ts(fitted(fm0), start = first_year), col = 3)
lines(ts(fitted(fm1), start = first_year), col = 2)
lines(ts(fitted(fm2), start = first_year), col = 4)

### Use ANOVA to compare models
anova(fm0, fm1, fm2)
```



