---
title: "Threshold"
author: "Worked example 10.1"
output: 
  tufte::tufte_html:
    tufte_features: ["fonts"]
    toc: true
    toc_depth: 2
    number_sections: true
    css: style.css
    fig_caption: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw(base_size = 10))
library(viridisLite)
```


# Loading the Data

Daily data from River Ngaruroro at Kuripapango (NZ) are used as an example. 


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(hydroDrought)

ngaruroro <- intl %>%
  filter(river == "Ngaruroro") %>%
  select(data) %>%
  unnest(data) %>%
  mutate(
    year = water_year(time, origin = "-09-01")
  )
```


Threshold calculations always remove missing values. 

Daily varying thresholds are more complicated because leap years need to be dealt with. 


# Constant tresholds 

The aggregating function can be any R function returning a single value like `mean()`, `quantile()`, `prob()`, ...

```{r}
const_threshold(ngaruroro, fun = lfquantile, exc.freq = 0.8)
const_threshold(ngaruroro, fun = median)
```


# Varying thresholds

Can vary by: day, week, month, season, year


```{r}
# specify the start of the season
# year will be ignored, only month and day are relevant
starts <- c("1971-02-24", "1971-06-03", "1971-10-22")
starts <- c("-02-24", "-06-03", "-10-22")
ngaruroro %>%
  var_threshold(vary.by = "season", fun = lfquantile, exc.freq = 0.8, start = starts)


# it is better to name the seasons
starts <- c("summer" = "-09-01", "winter" = "-02-01")
var_threshold(ngaruroro, vary.by = "season", 
              fun = lfquantile, exc.freq = 0.8, start = starts)
```


# Threshold for a subset of the data
Calculate threshold (Q70) for given years (eg. a subset of the whole time series). 
```{r}
threshold <- ngaruroro %>%
  filter(year >= 1970, year < 1990) %>%
  var_threshold(
    vary.by = "month", 
    fun = lfquantile, exc.freq = 0.7
  )

threshold
```


Use the threshold for the whole dataset. First we have to append a new column containing the groups we have calculated the threshold for (in our case months). Then the threshold will be joined based on that column. 
```{r}
# hydrological year starts in September (here only relevant for plotting)
grouped <- ngaruroro %>%
  mutate(
    month = group_ts(time = time, by = "month", origin = "-09-01")
  ) 

grouped

ng <- grouped %>%
  left_join(threshold, by = "month") %>%
  mutate(
    under.drought = discharge < threshold
  )

ng
```



Demonstrate monthly threshold for three years not used in threshold calculation. 
```{r}
ng %>%
  filter(year %in% 1964:1966) %>%
  ggplot() + 
  geom_line(aes(x = time, y = discharge)) + 
  geom_step(aes(x = time, y = threshold), 
            linetype = "dashed", size = 0.2, col = "red") + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_log10() +
  facet_wrap(~year, scales = "free_x", ncol = 1, strip.position = "right")
```


# Daily varying threshold with smoothing

```{r}
threshold <- ngaruroro %>%
  mutate(
    discharge = moving_average(discharge, n = 31, sides = "center")
  ) %>%
  filter(year >= 1970, year < 1990) %>%
  var_threshold(
    vary.by = "day", start = "-09-01", 
    fun = lfquantile, exc.freq = 0.8
  ) 

threshold

ng <- ngaruroro %>%
  mutate(
    day = group_ts(time = time, by = "day", origin = "-09-01")
  ) %>%
  left_join(threshold, by = "day") %>%
  print()
```

```{r, echo=FALSE}
p1 <- ng %>%
  filter(is.finite(discharge)) %>%
  plot_coverage() + 
  theme(legend.position = "top")

# p1 <- benchmark %>%
#   mutate(
#     day = group_ts(time = time, by = "day")
#   ) %>%
#   left_join(threshold, by = "day") %>%
#   
#   ggplot(aes(x = day, y = year, fill = threshold)) + 
#   geom_tile() + 
#   scale_fill_viridis_c() + 
#   scale_x_date(date_breaks = "1 months", date_labels = "%b") + 
#   theme(legend.position = "top", 
#         axis.title.x = element_blank())

p2 <- ggplot(data = threshold, aes(x = day, y = threshold)) + 
  geom_step() + 
  scale_x_date(date_breaks = "1 months", date_labels = "%b", 
               expand = expansion()) + 
  theme(axis.title.x = element_blank())

cowplot::plot_grid(p1, p2, align = "v", axis = c("l", "r"), 
                   ncol = 1, rel_heights = c(5, 2))
```



# Fast Track
The functions `var_threshold()` and `const_threshold()` both accept an argument `append = TRUE` to facilitate the computation of the threshold using the entire time series and appending it. 

```{r}
ngaruroro %>%
  var_threshold(vary.by = "month", 
                fun = lfquantile, exc.freq = 0.8, 
                append = TRUE) 
```

