---
title: "Human Influences"
author: "Worked example 10.1"
output:
    tufte::tufte_html:
    tufte_features: ["fonts"]
css: style.css
editor_options:
    chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw(base_size = 10))
library(viridisLite)
library(kableExtra)
```

^[Tobias: What will be the name of the Worked Example?]


```{r, echo = TRUE, message=FALSE, warning=FALSE}
library(hydroDrought)
# Let's pretend the data is already inside the package...
origin <- "-09-01"
guadiana <- read_tsv("../../data/Guadiana_Delay_LogEff_including_Psim.txt") %>%
    select(time = "date.yyyymmdd",  Qsim, Qobs)  %>%
    append_group(by = "day", start = origin)
```


```{r, message=FALSE, warning=FALSE}
library(hydroDrought)
guadiana
```

If we want to quantify the human influence on hydrological drought by comparing two time series, one with and one without this human influence, we want to use the threshold of the **benchmark** time series to calculate droughts in both the **human-influenced** and **benchmark** time series. These are the steps that we will discuss in this Worked Example:

1) Calculate the threshold from the **benchmark** time series

2) Calculate drought characteristics for the **benchmark** time series, using the **benchmark** threshold 

3) Calculate drought characteristics for the **human-influenced** time series, using the **benchmark** threshold

4) Compare drought characteristics between the **benchmark** & **human-influenced** time series


# 1) Threshold calculation

Calculate the threshold from the **benchmark** time series.

Load the **benchmark** time series. Note that these can be calculated from a paired catchment analysis (Sect. XX), an upstream-downstream comparison (Sect. XX), model naturalisation (Sect. XX), or pre-post disturbance analysis (Sect. XX). As an example we here use the Upper-Guadiana data^[Tobias: What ist the start of the low-flow season? Or can we simply use calendar years? ]. The **benchmark** time series for this catchment are modelled as described in Sect. XX.
Choose the threshold. 

```{r}
natural <- guadiana %>%
    select(time, day, year, discharge = Qsim)
```

Here we are taking the example of a daily varying threshold, smoothed by a 30-day moving average. We are using the period 1960-2000 to calculate the threshold. ^[Tobias: Which aggregation function should we use? $Q_{95}$]

```{r}
threshold <- natural %>%
    mutate(discharge = moving_average(discharge, n = 30, sides = "center")) %>%
    filter(year >= 1960 & year <= 2000) %>%
    var_threshold(vary.by = "day", start = origin, 
                  fun = lfquantile, exc.freq = 0.95) %>%
    print()
```


```{r, echo=FALSE}
p1 <- natural %>%
    left_join(threshold, by = "day") %>%
    filter(is.finite(discharge)) %>%
    ggplot(aes(x = day, y = year, fill = threshold)) + 
    geom_tile() + 
    scale_fill_viridis_c() + 
    scale_x_date(date_breaks = "1 months", date_labels = "%b") + 
    theme(legend.position = "top", 
          axis.title.x = element_blank())

p2 <- ggplot(data = threshold, aes(x = day, y = threshold)) + 
    geom_step() + 
    scale_x_date(date_breaks = "1 months", date_labels = "%b") + 
    theme(axis.title.x = element_blank())

cowplot::plot_grid(p1, p2, align = "v", axis = c("l", "r"), 
                   ncol = 1, rel_heights = c(5, 2))
```


# 2) Drought characteristics (benchmark time series/ benchmark threshold)
Calculate drought characteristics for the **benchmark** time series, using the **benchmark** threshold. ^[Tobias: Should we use pooling?]

Extent the **benchmark** threshold calculated in step 1 to the period of the **benchmark** time series. Note that this can be a different time period than was used to calculate the **benchmark** threshold (for example, for Upper-Guadiana, we are using the period 1981-2000).

```{r}
# thres.bench2 <- natural %>%
#     mutate(discharge = moving_average(discharge, n = 30, sides = "center")) %>%
#     filter(year >= 1981 & year <= 2000) %>%
#     var_threshold(vary.by = "day", fun = lfquantile, exc.freq = 0.95)
```


Calculate the drought characteristics you want to compare. For the Upper-Guadiana, these would be the **benchmark** drought characteristics: ^[Tobias: filter minor droughts?]

```{r}
dc.benchmark <- natural %>%
    filter(year >= 1981 & year <= 2000) %>%
    drought_events(threshold = threshold, pooling = "seq") 

kable(dc.benchmark)
```


# 3) Drought characteristics (human-influenced time series/ benchmark threshold)    
Calculate drought characteristics for the **human-influenced** time series, using the **benchmark** threshold

Load the **human-influenced** time series. In this case we are using the observed time series from the Upper-Guadiana catchment.
Extent the **benchmark** threshold calculated in step 1 to the period of the **human-influenced** time series. Note that this can be a different time period than was used to calculate the **benchmark** threshold (for example, for Upper-Guadiana, we are using the period 1981-2000).

```{r}
influenced <- guadiana %>%
    select(time, day, year, discharge = Qobs) %>%
    filter(year >= 1991 & year <= 2000)

```


```{r, warning=FALSE, fig.width=8, fig.height=10}
infl <- influenced %>%
    left_join(threshold, by = "day")

infl %>%
    ggplot() + 
    geom_line(aes(x = time, y = discharge)) + 
    geom_step(aes(x = time, y = threshold), 
              linetype = "dashed", size = 0.2, col = "red") + 
    scale_x_date(date_breaks = "1 month", date_labels = "%b") +
    scale_y_log10() +
    facet_wrap(~year, scales = "free_x", ncol = 1, strip.position = "right")
```


Calculate the drought characteristics you want to compare.
For the Upper-Guadiana, these would be the **human-influenced** drought characteristics:

```{r}
dc.influenced <- influenced %>%
    drought_events(threshold = threshold, pooling = "sequent-peak") 

kable(dc.influenced)
```



# 4) Comparison of drought characteristics    

Compare drought characteristics between the **benchmark** & **human-influenced** time series

```{r}
summarize_dc <- function(x) {
    c("mean.duration" = mean(x$duration), 
      "mean.deficit" = mean(x$volume))
}
```

Calculate the percentage difference between the **benchmark** and **human-influenced** drought characteristics.

$$\Delta DC = \frac{DCHI - DCBM}{DCBM} \cdot 100$$
where $\Delta DC$ is the percentage change in drought characteristics ($DC$) between the **human-influenced** ($DCHI$) and **benchmark** ($DCBM$) time series.
For the Upper-Guadiana, these would be the differences in drought characteristics:

```{r}
dc.hi <- summarize_dc(dc.influenced)
print(dc.hi)

dc.bm <- summarize_dc(dc.benchmark)
print(dc.bm)

(dc.hi - dc.bm) / dc.bm * 100
```

