---
title: "Threshold"
author: "Worked example 10.1"
output: 
  tufte::tufte_html:
    tufte_features: ["fonts"]
    css: style.css
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw(base_size = 10))
library(kableExtra)
```


# Loading the Data

Daily data from River Ngaruroro at Kuripapango (NZ) are used as an example. 


```{r, message=FALSE, warning=FALSE}
library(hydroDrought)

ngaruroro <- filter(international, River == "Ngaruroro") %>%
    select(discharge) %>%
    unnest()
```


Threshold calculations always remove missing values. 

Daily varying thresholds are more complicated because leap years need to be dealt with. 


## Constant tresholds 

The aggregating function can be any R function returning a single value like `mean()`, `quantile()`, `prob()`, ...

```{r}
const_threshold(ngaruroro, fun = lfquantile, exc.freq = 0.8)
const_threshold(ngaruroro, fun = median)
```


## Varying thresholds

Can vary by: day, week, month, season, year


```{r}
# specify the start of the season
# year will be ignored, only month and day are relevant
starts <- c("1971-02-24", "1971-06-03", "1971-10-22")
starts <- c("-02-24", "-06-03", "-10-22")
ngaruroro %>%
    var_threshold(vary.by = "season", fun = lfquantile, exc.freq = 0.8, start = starts)


# it is better to name the seasons
starts <- c("summer" = "-09-01", "winter" = "-02-01")
var_threshold(ngaruroro, vary.by = "season", 
              fun = lfquantile, exc.freq = 0.8, start = starts)
```


## Subset
Calculate threshold (Q70) for given years (eg. a subset of the whole time series). 
```{r}
threshold <- ngaruroro %>%
    filter(time >= as.Date("1970-01-01") & time < as.Date("1990-01-01")) %>%
    var_threshold(vary.by = "month", fun = lfquantile, exc.freq = 0.7)

threshold
```


Use the threshold for the whole dataset. First we have to append a new column containing the groups we have calculated the threshold for (in our case months). Then the threshold will be joined based on that column. 
```{r}
# hydrological year starts in September (only relevat for plotting)
grouped <- ngaruroro %>%
    append_group(by = "month", start = "-09-01")

grouped

ng <- grouped %>%
    left_join(threshold, by = "month") %>%
    mutate(under.drought = discharge < threshold)

ng
```



Demonstarte montly threshold for three years not used in threshold calculation. 
```{r}
ng %>%
    filter(year %in% 1964:1966) %>%
    ggplot() + 
    geom_line(aes(x = time, y = discharge)) + 
    geom_step(aes(x = time, y = threshold), 
              linetype = "dashed", size = 0.2, col = "red") + 
    scale_x_date(date_breaks = "1 month", date_labels = "%b") +
    scale_y_log10() +
    facet_wrap(~year, scales = "free_x", ncol = 1, strip.position = "right")
```

