---
title: "Flow Duration Curve"
author: "Worked Example 5.1"
output: 
  tufte::tufte_html:
    tufte_features: ["fonts"]
    toc: true
    toc_depth: 2
    number_sections: true
    css: style.css
    fig_caption: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw(base_size = 10))
library(kableExtra)
```


# Loading the Data

In this example we are going to use river flow data from the river Ngaruroro which is part of the `international` dataset in the package **hydroDrought**. This dataset becomes accessible by loading the **hydroDrought** package. 
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(hydroDrought)
ngaruroro <- intl %>%
  filter(river == "Ngaruroro") %>%
  select(data) %>%
  unnest(data) %>%
  print()
```

All years of daily data from River Ngaruroro at Kuripapango (NZ) are used here to construct a flow duration curve based on a daily time step, $\Delta t = 1$ day. The total number of $\Delta t$ intervals is then $N = `r nrow(ngaruroro)`$ days. Table 5.1 lists the first seven flow values. The first two columns show the date and the corresponding riverflow value, $Q$.




# Calculation of the FDC

The flow duration curve is constructed following the calculation steps as given in the right part of the table:

- The rank, $i$, of each value is calculated (using the `rank()` function), which means that if the list is sorted, the rank will be its position. Here the series is sorted in descending order and the $i^{th}$ largest value has rank $i$ (i.e. the largest value has rank 1).

- The exceedance frequency, $EF_{Q_i}$ is calculated as: 
$$EF_{Q_i}  = \frac{i} {N}$$ 
which gives an estimate of the empirical exceedance frequency of the $i^{th}$ largest event. $EF_{Q_i}$ designates here the observed frequency when the flow, $Q$, is larger than the flow value with rank $i$, $Q_i$ .

```{r}
exceedance_frequency <- function(flow)
{
  # current rank
  i <- rank(-flow, ties.method = "min", na.last = "keep") 
  
  # largest rank in sample (= number of non-missing values)
  N <- length(na.omit(flow))
  
  # the exceedance frequency can be seen as the relative rank
  return(i / N)
}

ngaruroro <- ngaruroro %>%
  mutate(
    rank = rank(-discharge, ties.method = "min"), 
    freq.exc = exceedance_frequency(discharge)
  )
```


```{r, echo=FALSE}
ngaruroro %>%
  mutate_if(is.numeric, round, 3) %>%
  select(Date = time, Flow = discharge, Rank = rank, `Exceedance Frequency` = freq.exc) %>%
  head(7) %>%
  kable(caption = "Table 5.1 Calculation of a daily flow duration curve for River Ngaruroro at Kuripapango, NZ") %>%
  add_header_above(c(" " = 1, "$Q$ in $m^3s^{-1}$" = 1, "$i$" = 1, "$EF_Q$" = 1), 
                   align = "right")
```




# Tabulation of the FDC

- Corresponding values of riverflow ($Q$ in m<sup>3</sup>s<sup>-1</sup>) and exceedance frequency ($EF_{Q_i}$ in %) are tabulated^[Tobias: Tabulation was done implicitly in the sections before. ].

- The two columns are sorted by $EF_{Q_i}$^[Tobias: Sorting is not required. I suggest to delete the section "Tabulation of the FDC" entirely.].

# Plot of the FDC
The sorted table columns are then plotted (Figure 5.2). The ordinate axis is here logarithmic. 

```{r, fig.cap="Figure 5.2 Flow duration curve for River Ngaruroro at Kuripapango, NZ.", warning=FALSE, fig.margin = FALSE, message=FALSE}
library(scales)
ggplot(ngaruroro, aes(x = freq.exc, y = discharge)) +
  geom_line() +
  scale_y_log10() +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Exceedance Frequency",
       y = expression(paste("Flow (", m^{3}, s^{-1}, ")")))
```



# Selected exceedance values

```{r, echo=FALSE}
q90 <- ngaruroro %>%
  filter(freq.exc <= 0.9) %>%
  arrange(desc(freq.exc)) %>%
  head(1) %>%
  pull(discharge)
```

Values for a particular frequency, for example the 90-percentile ($Q_{90}$), can be obtained as the value of $Q$ corresponding to the largest value of $EF_{Q_i}$ that is less than or equal to the value of $EF_{Q_i}$ sought for. A sample of corresponding values in this range is shown in Table 5.2, and the 90-percentile flow value is taken as `r q90`<i>&nbsp;</i>m<sup>3</sup>s<sup>-1</sup>. Alternatively, in case of large differences between successive values, a linear interpolation can be used. 

```{r include=FALSE}
tbl <- ngaruroro %>%
  filter(freq.exc >= 0.8999 & freq.exc <= 0.9002) %>%
  arrange(freq.exc)
```


```{r echo=FALSE}
tbl  %>%
  mutate(discharge = round(discharge, 3), 
         freq.exc = round(freq.exc, 5)) %>%
  select(Date = time, Flow = discharge, Rank = rank, `Exceedance Frequency` = freq.exc) %>%
  kable(caption = "Table 5.2 An extract of values corresponding to $Q_{90}$.") %>%
  add_header_above(c(" " = 1, "$Q$ in ${m^3s^{-1}}$" = 1, "$i$" = 1, "$EF_Q$" = 1), 
                   align = "right")
```

# Fast-Track

The function `lfquantile()` calculates low-flow/riverflow quantiles directly. The exact (interpolated) values for $Q_{95}$, $Q_{90}$ and $Q_{80}$ would be: 

```{r}
lfquantile(ngaruroro$discharge, exc.freq = c(0.95, 0.9, 0.8))
```


# Standardized Flow Duration Curves

Comparing FDCs from different catchments requires standardization eg. dividing the discharges by the catchment area, the median or the mean discharge. 
```{r, warning=FALSE, message=FALSE}
lambournRay <- intl %>%
  filter(river %in% c("Lambourn", "Ray")) %>%
  select(river, area = catchment, data)

lambournRay <- lambournRay %>%
  mutate(
    data = map2(data, area, ~mutate(.x, rel.discharge = discharge * 1000 / .y))
  ) %>%
  print()

fdc <- lambournRay %>%
  select(river, data) %>%
  unnest(cols = data) %>%
  group_by(river) %>%
  mutate(freq.exc = exceedance_frequency(rel.discharge)) %>%
  print()


ggplot(fdc, mapping = aes(x = freq.exc, y = rel.discharge, 
                          size = river)) +
  geom_step(direction = "vh") +
  scale_size_manual(values = c(0.1, 0.75)) +
  scale_y_log10(limits = c(0.01, NA), 
                labels = number_format(accuracy = 0.01)) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Exceedance Frequency",
       y = expression(paste("Flow (", l, s^{-1}, km^{-2}, ")"))) 
```

