---
title: "Mean annual minimum flow"
author: "Worked example 5.2"
output: 
   tufte::tufte_html:
    tufte_features: ["fonts"]
    toc: true
    toc_depth: 2
    number_sections: true
    css: style.css
    fig_caption: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_bw(base_size = 10))
library(kableExtra)
```


# Loading the Data

In this example we are again going to use river flow data from the river Ngaruroro of the `international` dataset in the package **hydroDrought**. 
Ten years of daily data are used as an example, as in Worked Example 5.1, to estimate mean annual minimum of the $n$-day average flow for $n$ equal to 1, 7 and 30 days. For this station the lowest flows are observed around the turn of the calendar year. Therefore the annual minima are selected from **hydrological years** starting 1 September and ending 31 August. Table 5.4 lists the first flow values. 
<!-- The first two columns show the date and the corresponding flow value, Q. -->

In order to calculate the mean annual minimum each observation will be attributed to a hydrological year according to the date of the observation using the function `water_year()` which appends an additional column named `year` to the dataset. 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(hydroDrought)

# attribute each observation to the correct hydrological year
# and select only the hydrological years between 1990 und 2000
ngaruroro <- intl %>%
  filter(river == "Ngaruroro") %>%
  select(data) %>%
  unnest(data) %>%
  mutate(
    year = water_year(time, start = "-09-01")
   ) %>%
  filter(year >= 1990, year <= 1999)

smoothed <- ngaruroro %>%
  mutate(
    MA1 = moving_average(discharge, n = 1),
    MA7 = moving_average(discharge, n = 7),
    MA30 = moving_average(discharge, n = 30)
  )
```

```{r, echo=FALSE}
num_in_brackets <- function(x) sub("(\\d+)", "(\\1)", x = x)
library(kableExtra)
smoothed %>%
  select(-discharge, -year) %>%
  head(31) %>%
  mutate_if(is.numeric, round, digits = 3) %>%
  rename_all(.funs = num_in_brackets) %>%
  kable(caption = "Table 5.4 Calculation of n-day average flow (unit: in m<sup>3</sup>s<sup>-1</sup>), River Ngaruroro at Kuripapango, NZ. ") %>%
  add_header_above(c("Data, 10-year series" = 2, "Moving average calculation" = 2), 
                   align = "right")
```


```{r}
# compute the annual minima
am <- smoothed %>%
  select(-discharge, -time) %>%
  group_by(year) %>%
  summarise_all(min, na.rm = TRUE)

# average the annual minima to get the mean annual minima
mam <- am %>% 
  select(-year) %>%
  summarise_all(mean)
```

```{r, echo=FALSE}
area <- filter(international, River == "Ngaruroro") %>%
  pull(Area)
tbl <- tibble(n = gsub("\\D", "", names(mam)),
              discharge = unlist(mam), 
              rel.discharge = discharge * 1000 / area) 
mam  %>%
  mutate_if(is.numeric, round, digits = 3) %>%
  rename_all(.funs = num_in_brackets) %>%
  kable(caption = "Table 5.5 $MAM$(n-day), $n=1$ day, 7 days and 30 days") %>%
  add_header_above(c("$Q$ in ${m^3s^{-1}}$" = 3), align = "center")
```

# Fast Track

```{r}
ngaruroro %>% 
  select(discharge, time) %>%
  summarise(
    `MAM(1)` = mean_annual_minimum(discharge, time, start = "-09-01", n = 1),
    `MAM(7)` = mean_annual_minimum(discharge, time, start = "-09-01", n = 7),
    `MAM(30)` = mean_annual_minimum(discharge, time, start = "-09-01", n = 30)
  )
```

