# Chapter 6

## Worked Example 6.1: Low flow frequency analysis

### Loading the Data

River Ngaruroro at Kuripapango in New Zealand (Table 4.3), has been
selected for frequency analysis of annual minimum 1-day values, $AM(1)$,
using the Weibull (WEI) distribution. In mid-latitudes in the Northern
Hemisphere, the calendar year is often used to dplyr::select the annual
minimum flows. This is a suitable period for the selection of independent
events as the drought or low flow period commonly occurs during the
summer months. In the Southern Hemisphere the low flow season occurs at
the opposite time of the year, and for Ngaruroro the lowest flows are
typically found in the period November to May. As a result $AM(1)$ flows
were selected for a hydrological year starting at 1 September.


```{r}
library(tidyverse)
library(hydroDrought)

ngaruroro <- international %>%
  filter(river == "Ngaruroro") %>%
  dplyr::select(data) %>%
  unnest(data) %>%
  # linear interpolation of periods of missing values 
  # with a duration of up to 15 days
  sanitize_ts(approx.missing = 15) %>%
   mutate(
    year = water_year(time, origin = "-09-01")
    ) %>%
  print()
```


```{r}
# remove incomplete years
incomplete <- ngaruroro %>%
  filter(!is.na(discharge)) %>%
  pull(time) %>%
  coverage_yearly(origin = "-09-01") %>%
  filter(days.missing > 0) %>%
  print()

ngaruroro <- ngaruroro %>%
    anti_join(incomplete, by = "year")
```


```{r}
# derivation of the annual minima
am <- annual_minima(
  discharge = ngaruroro$discharge, 
  time = ngaruroro$time, 
  origin = "-09-01"
  ) %>%
  rename(flow = am) %>%
  arrange(flow) %>%
  print()
```

```{r, echo=FALSE}
fig.cap <- "Figure 6.9 Time series of annual minimum 1-day flow, AM(1), for River Ngaruroro at
Kuripapango (NZ)."
```
```{r, echo=FALSE, fig.cap=fig.cap}
ggplot(am, aes(x = year, y = flow)) + 
  geom_point() + 
  # geom_smooth(method = "loess", formula = "y ~ x", span = 1, se = FALSE,
  #             col = "black", size = 0.2, linetype = "dashed") + 
  labs(y = expression(paste("Annual minimum flow (", m^{3}, s^{-1}, ")"))) +
  scale_y_continuous(limits = c(0, NA)) + 
  theme(axis.title.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

Ngaruroro has a flashy river regime, but has no observed zero flows. The  flows 
are considered to come from the same population as only a minor part of the 
catchment is influenced by snow in winter. The observations (daily flow series) 
cover the period `r trimws(format(min(ngaruroro$time), format = "%e %B %Y"))` to 
`r trimws(format(max(ngaruroro$time), format = "%e %B %Y"))`. With the omission 
of `r nrow(incomplete)` 
years (`r hydroDrought:::paste_with_and(incomplete$year)`) with missing data, a 
total of `r nrow(am)` annual minimum values results. The procedure used for 
interpolation of missing data is the same as in WE5.4. A histogram of the values 
is shown in Figure 6.3 (upper left). To test the assumption of stationarity, 
the  values are plotted against time in Figure 6.9. No trend can be detected 
in the series and the data are therefore assumed to fulfil the requirement of 
independent and identically distributed data (iid).


### Derivation of an empirical distribution function

```{r, empirical, echo=FALSE}
empirical <- am %>%
  mutate(
    rank = rank(flow),
    prob.emp = rank / (n() + 1),
    rp.emp = 1 / prob.emp
  ) %>%
  arrange(rank)  
```

a) The $x$ values, AM(1), are sorted in ascending order and the rank of
each value is calculated. The smallest value equals `r min(am$flow)` in m<sup>3</sup>s<sup>-1</sup> and is given rank 1.

b) The non-exceedance probability, $F(x) = p$, (`'prob.emp'` in the tibble below) is calculated for each $x$
value using the Weibull plotting position formula $p_i = i / (n + 1)$ (Equation 6.6).

c) A probability plot for the $AM(1)$ values is obtained by plotting the
flow values against $F(x)$ as demonstrated in Figure 6.10 (upper). A
staircase pattern of several nearly equal values is observed. (Note that
the $AM(1)$ values now are plotted on the $y$-axis as compared to
Figure 6.3, lower).



d) The return period (`'rp.emp'`) of the smallest event can be calculated following
Equation 6.5. The non-exceedance frequency, $F(x)$, of the smallest
event equals `r round(empirical$prob.emp[1], 3)` according to step 2(b), which gives a return period
of `r empirical$rp.emp[1]`  years.

```{r, empirical, eval=TRUE, echo=TRUE}
```
```{r}
print(empirical)
```

### Fitting a distribution  
Fitting the two-parameter Weibull (WEI) distribution function
(Section A6.1.4) (assuming the location parameter $\xi$ is zero) using the method of L-moments. 
^[A direct estimation of the Weibull parameters is shown in the [Fast Track](#ft) section. ]


a) The first two L-moments ($\hat\lambda_1$ and $\hat\lambda_2$) are estimated 
based on time series of the annual minima `am$flow`. These so called sample 
L-moments are calculated with the function `samlmu()` from the package **lmom**.

    ```{r}
library(lmom)
moments <- samlmu(am$flow)
moments[1:2] 
    ```

b) The parameter estimates of the WEI distribution (scale parameter $\alpha$, 
shape parameter $\kappa$) are obtained following Equation A6.144:

    $$\hat\alpha = \frac{\hat\lambda_1}{\Gamma\left( 1 + \frac{1}{\hat\kappa }\right)}$$ 

    $$\hat\kappa = \frac{-\ln(2)}{\ln\left( 1 - \frac{\hat\lambda_2}{\hat\lambda_1}\right)}$$ 


    ```{r}
kappa <- -log(2) / log(1 - unname(moments["l_2"] / moments["l_1"]))
alpha <- unname(moments["l_1"]) / gamma(1 + 1/kappa)
parameter <- c(location = 0, scale = alpha, shape = kappa)
parameter
    ```

c) $F(x)$ for the WEI distribution is obtained following Equation A6.1.33
(recalling that the location parameter $\xi$ is set to zero):

    $$ F(x) = 1 - \exp\left[-\left(\frac{x}{\alpha}\right)^\kappa\right]$$
    
    ```{r}
fitted <- empirical %>%
  dplyr::select(year, flow, prob.emp) %>%
  mutate(
    prob.wei = 1 - exp(-(flow / alpha)^kappa),
    rp.wei = 1 / prob.wei
  ) 
    ```

d) The data can be plotted in a probability plot and compared to the
empirical quantiles in step 2(b) (Figure 6.10, upper). The plot shows
that the WEI distribution is well adjusted to the low flow extreme
values, whereas deviation in the upper range may suggest that the
three highest values do not belong to the low flow population
(Section 6.3.3). In Figure 6.10 (lower) the empirical quantiles are
plotted against the estimated distribution quantiles (QQ-plot). The
points should be close to the unit diagonal if the data fit the WEI
distribution well. The use of a plotting position implies that the
ordered sample is plotted in regularly spaced positions. As
demonstrated in the figure, the observed jumps in $AM(1)$ values are
reflected in the estimated Weibull quantiles, but not in the empirical
quantiles.

```{r, echo=FALSE}
fig.cap <- "Figure 6.10 (upper): Estimated quantiles for annual minimum 1-day flow, AM(1), for River
Ngaruroro at Kuripapango (NZ); probability plot showing the fit for the Weibull distribution
(curve) to the sample data."
```
```{r, echo=FALSE, fig.cap=fig.cap}
# quant.theo = qweibull(empirical, shape = kappa, scale = alpha),
# quant.theo2 = exp(log(-log(1-empirical))/kappa + log(alpha))

# tocheck: we could estimate the fitted distribution much smoother... 
ggplot(fitted, aes(prob.emp, flow)) + 
  geom_point() + 
  geom_line(aes(x = prob.wei)) + 
  scale_y_continuous(limits = c(0, NA)) + 
  labs(x = expression(paste("F(", x, ")")),
       y = expression(paste("Annual minimum flow (", m^{3}, s^{-1}, ")")))


```



```{r, echo=FALSE}
fig.cap <- "Figure 6.10 (lower): QQ-plot of empirical quantiles versus Weibull quantiles"
```
```{r, echo=FALSE, fig.cap=fig.cap}
# todo: in the old book, these are not quantiles! 
ggplot(fitted, aes(x = prob.emp, y = prob.wei)) + 
  geom_point() + 
  labs(x = "Empirical quantiles", y = "Theoretical Quantiles") + 
  geom_abline(slope = 1, intercept = 0, size = 0.2, linetype = "dashed") + 
  coord_fixed()
```



### Estimation of T-year events

e) The non-exceedance frequency of the smallest value equals 
`r round(fitted$prob.wei[1], 4)` (Equation A6.1.33), which gives a return period of 
`r round(fitted$rp.wei[1], 1)` years (Equation 6.5). 
This is `r round(empirical$rp.emp[1] - fitted$rp.wei[1], 1)` 
years less that the empirical estimate derived in step 2(d).

    ```{r}
fitted
    ```


f) The 50 and 100-year events can be estimated from Equation A6.1.34:
$$ x_p = \xi + \alpha \left[-\ln( 1 - p ) \right]^{1/\kappa} $$

```{r, quantiles, echo=FALSE}
quantile_weibull <- function(prob, location = 0, scale, shape) {
  location + scale * (-log(1 - prob))^(1/shape)
}

return.period <- c(50, 100)
q <- quantile_weibull(prob = 1/return.period, scale = alpha, shape = kappa) 
```

which gives: $\hat x_{50} = `r format(q[1], digits = 2, nsmall = 2)`$m<sup>3</sup>s<sup>-1</sup> 
and $\hat x_{100} = `r format(q[2], digits = 2, nsmall = 2)`$m<sup>3</sup>s<sup>-1</sup>. 


```{r, quantiles, echo=TRUE}
```
```{r}
round(q, 2)
```



### Fast Track {#ft}

The function `samlmu()` from the package **lmom** computes the sample L-moments 
for a given sample. These sample L-moments are independent of the distribution one
wants to fit. They have can be transferred into the parameters of the desired
distribution; e.g. the function `pelwei()` will return the location, scale and 
shape parameters for a Weibull distribution. By default a three parameter Weibull
distribution is used. To obtain L-moment estimates of the two parameter Weibull 
distribution (where the lower bound and location parameter $\xi$ is set to zero) 
set the argument `bound = 0`.
```{r}
library(lmom)
lmom <- samlmu(am$flow)
par.wei <- pelwei(lmom, bound = 0)
par.wei
```

Quantiles for given probabilities can be calculated for many suitable 
distributions by the corresponding quantile function `quawei()`. 

```{r}
quawei(f = c(0.02, 0.01), para = par.wei) %>%
  round(digits = 2)
```

### Probability plot using the exponential reduced variate

The two-parameter EXP (Equation A6.1.64) and the WEI distribution are compared
using the exponential reduced variable (Box 6.4). 

```{r}
fitted.fast <- am %>%
  mutate(
    prob.emp = rank(flow) / (n() + 1),
    prob.wei = cdfwei(x = flow, para = pelwei(lmom, bound = 0)),
    prob.exp = cdfexp(x = flow, para = pelexp(lmom))
  ) %>% 
  arrange(flow) %>%
  print()
```

a) By substituting $y$ into the expression for $F(x)$, $y$ can be
expressed as $-\ln(1 - F(x))$ (Equation B6.4.6). The parameters of the
EXP distribution are estimated using L-moments (Equation A6.1.72)
and used to calculate $y$ and subsequently, $F(x)$.

b) The non-exceedance probability $F(x)$ for the observations is
determined using the Weibull plotting position formula
(Equation 6.6).

c) The $AM(1)$ values are plotted against the reduced variate in
Figure 6.11. A reduced variate of $0.02$ corresponds to a return period
of $50.5$ years for minimum values (Equation 6.5). The data will plot as
a straight line given they follow the EXP distribution. Again it is
demonstrated that the WEI distribution fits the extreme low flow
values well, and also the upper range apart from the three largest
values. The two-parameter EXP distribution is less suited to model the
sample. Alternatively, $-\ln(X)$ could be plotted on a Gumbel
probability paper, and a straight line would result provided the data
fitted the WEI distribution.

```{r, echo=FALSE}
fig.cap <- "Figure 6.11: The annual minimum 1-day flow, AM(1), plotted against the reduced variate, y, of the EXP distribution for River Ngaruroro at Kuripapango (NZ); the observations (marked as points), the two-parameter EXP distribution (continous line) and the two-parameter WEI distribution (dashed line)."
```


```{r, echo=FALSE, warning=FALSE, fig.cap=fig.cap}
reduced_var <- function (x) -log(1 - x)

d <- fitted.fast %>%
  dplyr::select(year, flow, prob.wei, prob.exp) %>%
  pivot_longer(c(prob.wei, prob.exp), 
               names_to = "Distribution", values_to = "probability") %>%
  mutate(Distribution = c(prob.wei = "Weibull", prob.exp = "Exponential")[Distribution])

ggplot(empirical, aes(x = reduced_var(prob.emp), flow)) + 
  geom_point() + 
  geom_line(data = d, 
            aes(x = reduced_var(probability), linetype = Distribution)) +
  scale_x_continuous(limits = c(0.01, 4)) +
  labs(x = expression(paste("Exponential reduced variate -log(", 1, " - F(", x, "))")),
       y = expression(paste("Annual minimum flow (", m^{3}, s^{-1}, ")")))
```


## Worked Example 6.2: Drought deficit frequency analysis

```{r, echo = FALSE}
library(tidyverse)
library(hydroDrought)
library(lubridate)

ngaruroro <- international %>%
  filter(river == "Ngaruroro") %>%
  dplyr::select(data) %>%
  unnest(data) 

ngaruroro <- ngaruroro %>%
  sanitize_ts(approx.missing = 14) %>%
  mutate(
    year = water_year(time, origin = "-09-01")
  ) 

coverage <- ngaruroro %>%
  filter(!is.na(discharge)) %>%
  pull(time) %>%
  coverage_yearly(origin = "-09-01")

incomplete <- coverage %>%
  filter(days.missing > 0) 

complete <- coverage %>%
  filter(days.missing == 0)

ngaruroro <- ngaruroro %>%
  anti_join(incomplete, by = "year")

q90 <- lfquantile(ngaruroro$discharge, exc.freq = 0.9) 

pooled1 <- ngaruroro %>%
  drought_events(
    threshold = q90, pooling = "inter-event", 
    pooling.pars = list(min.duration = 2, min.vol.ratio = Inf)
  ) %>%
  mutate(
    duration = as.double(dbt, units = "days"),
    volume = volume / 1e3
  )

threshold <- tibble(
  metric = c("duration", "volume"),
  threshold = c(5, round(max(pooled1$volume) * 0.005))
)

filtered <- pooled1 %>%
  filter(
    duration >= threshold %>% filter(metric == "duration") %>% pull(threshold),
    volume >= threshold %>% filter(metric == "volume") %>% pull(threshold) 
  ) %>%
  mutate(
    year = water_year(first.day, origin = "-09-01"),
  ) 

pooled <- filtered %>%
  dplyr::select(event, year, first.day, duration, volume) 

no_drought <- complete %>%
  anti_join(pooled, by = "year")
```

### Data
River Ngaruroro at Kuripapango in New Zealand, applied for frequency
analysis of annual minimum series in Worked Example 6.1, is here applied
for frequency analysis of drought deficit characteristics.

```{r}
library(tidyverse)
library(hydroDrought)

pooled
```

The pooled drought events obtained in Worked Example 5.4 (Section 5.4.1) can be
concieved as a partial duration series (PDS) 
because after removing `r nrow(pooled1) - nrow(filtered)` minor droughts only 
`r nrow(pooled)` events above and equal to the following thresholds remained
in the series. 

```{r}
threshold 
```


Two drought characteristics are analysed in the following: drought deficit 
volume and real drought duration as defined in Worked Example 5.4.



Drought events were selected for River Ngaruroro and details of the
selection criteria are given in Worked Example 5.4. The series cover the
period `r min(ngaruroro$year)` to `r max(ngaruroro$year)` and the start of the 
year is set to 1 September. 


A total of `r hydroDrought:::numbers_english(nrow(incomplete))` years were omitted from the series, 
`r hydroDrought:::paste_with_and(paste(incomplete$year, substr(incomplete$year + 1, 3L, 4L), sep = "/"))`, 
due to missing data. As a result `r nrow(complete)` years remained in the dataset. 

As only events below the $Q_{90}$
percentile are selected, it might happen that the flow never becomes less
than the threshold in a year (non-drought year). 

A total of `r hydroDrought:::numbers_english(length(no_drought))` out of the 
`r nrow(complete)` years with observations did not experience a drought 
(`r round(100 * nrow(no_drought) / nrow(complete), 1)`%). The PDS
series of drought deficit volume and real duration are plotted in
Figure 6.12. Less severe values are found in the second half of the
observation period for both deficit volume (upper) and duration (lower).
The data are still treated as one sample as the number of observations is
considered insufficient for a separate analysis of two periods. It should
further be noted that a similar trend towards less severe droughts is not as
pronounced for the $AM(1)$ values (Worked Example 6.1). This is likely a
result of the high base flow contribution in the catchment (Figure 6.6).




```{r, echo=FALSE}
fig.cap <- "Figure 6.12 PDS of drought deficit volume (upper) and real duration (lower) for River
Ngaruroro at Kuripapango (NZ)."
```
```{r, echo=FALSE, fig.cap=fig.cap}
library(cowplot)
p1 <- ggplot(pooled, aes(x = first.day, y = volume)) +
  geom_point(alpha = 0.4) + 
  scale_x_date(expand = expansion(mult = 0.01), date_minor_breaks = "1 year") +
  scale_y_continuous(limits = c(0, NA), 
                     expand = expansion(mult = c(0.01, 0.04))) + 
  labs(y = expression(paste("Deficit volume (", 10^{3}, m^{3}, ")"))) +
  theme(axis.title.x = element_blank())

p2 <- ggplot(pooled, aes(x = first.day, y = as.double(duration))) +
  geom_point(alpha = 0.4) + 
  scale_x_date(expand = expansion(mult = 0.01), date_minor_breaks = "1 year") +
  scale_y_continuous(limits = c(0, NA), 
                     expand = expansion(mult = c(0, 0.04))) +
  labs(y = "Real duration (days)") + 
  theme(axis.title.x = element_blank())

cowplot::plot_grid(p1, p2, align = "v", axis = "x", ncol = 1)
```



### Derivation of distribution function

in the [Journal of Statistical Software](https://www.jstatsoft.org/index.php/jss/article/view/v072i08/v72i08.pdf) there is an article describing the extremes package. 

Their approach is identical to Equation 6.11 in the first edition of the book. 

> "The quantiles of the GP df are easily obtained by setting Equation 5 equal to
$1 - p$ and inverting. However, the quantiles of the GP df cannot be as readily interpreted as
return levels because the data no longer derive from specific blocks of equal length. Instead,
an estimate of the probability of exceeding the threshold, e.g., called $\zeta_u$ , is required. Then,
the value $x_m$ that is exceeded on average once every $m$ observations (i.e., an estimated return
level) is "

$$ x_m = u + \frac{\sigma_u}{\xi} \left[ (m \zeta_u)^\xi - 1\right] $$
with $m$ being the return period and $\zeta_u$ the overall exceedance rate (= average number of exceedances 
per year). So the return period is just multiplied with the exceedance rate. 
$u$ is the location parameter of the GPA (= threshold), $\sigma_u$ is scale and $\xi$ is shape.
In our case $\zeta_u = \frac{\text{length of PDS}}{\text{record length in years}}$. 

This transformation will introduce return periods of less than a year (or negative 
probabilities) for values $P_{PDS} < 1 + \zeta_u$. Return periods for PDS of less 
than a year just imply that such an event occurs on average several times a year. 

$$P_{annual} = 1 - \frac{(1 - P_{PDS})}{\zeta_u}$$

I've seen PDS return periods < 1 for example in [Return Periods of Hydrological Events, Rojsberg 1976](https://iwaponline.com/hr/article/8/1/57/1454/Return-Periods-of-Hydrological-EventsPaper). I've also looked at his approach of relating PDS and AMS. But for return periods > 10 years quantiles of the AMS and PDS are practically identical, irrespective of the exceedance rate.   



Following Zelenhasic & Salvai (1987) an estimate of the non-exceedance
probability, $F(x)$, for the largest event in each time interval is in Nizowka
obtained by combining the distribution for the occurrence of events and
the distribution for the magnitudes of deficit volume or duration. Here a
time interval of one year is chosen. Subsequently the return period in
years for a given event can be calculated. The number of drought events
occurring in a time period t is commonly assumed to be Poisson
distributed (Equation 6.9) with parameter $\lambda t$. In Nizowka the binomial
Pascal distribution is offered along with the Poisson distribution as
described in ‘Background Information NIZOWKA’ (Software, CD). The
distribution that best fitted deficit volume was the Pascal distribution for
the number of droughts and the GP distribution for the deficits. For
duration the Pascal distribution was chosen along with the Log-Normal
distribution. The $F(x)$ is for drought deficit volume plotted in Figure 6.13
together with the observed values plotted using a plotting position. The
chosen distribution describes the data well, with the exception of some
values in the upper range. The maximum value is, however, satisfactorily
modelled.



```{r}
library(lmom)
rp <- c(10, 100, 200)
prob.annual <- 1 - c(1 / rp)

annual2pds <- function(prob.annual, exc.per.year) {
  1 - (1 - prob.annual)/exc.per.year
}

pds2annual <- function(prob.pds, exc.per.year) {
  1 - (1 - prob.pds) * exc.per.year
}

observations <- pooled %>% 
  dplyr::select(duration, volume) %>%
  mutate(across(everything(), as.numeric)) %>%
  pivot_longer(cols = everything(), names_to = "metric") %>%
  nest(data = value)  %>%
  left_join(threshold, by = "metric") 

fitted <- observations %>% 
  mutate(
    rate = map2_dbl(data, data, ~(nrow(.x) / nrow(complete))), 
    parameter = map2(data, threshold, ~pelgpa(samlmu(.x$value), bound = .y))
    ) %>%
  print()
```


```{r echo=FALSE}
reduced_var <- function (x) -log(1 - x)

o <- fitted %>%
  mutate(
    metric = factor(
      metric, levels = c("duration", "volume"),
      labels = c("Real duration\n(in days)", 
                 "Absolute volume\n(in 1000 m³)"))
  ) %>%
  unnest(data) %>%
  group_by(metric) %>%
  mutate(prob.annual = pds2annual(rank(value, ties.method = "random") / (n() + 1), rate[1])) 

f <- fitted %>%
  mutate(
    quafun = map2(parameter, rate, ~tibble(
      prob.pds = seq(0, 0.998, 0.001), 
      prob.annual = pds2annual(prob.pds, .y),
      quant = quagpa(prob.pds, .x)
    )),
    metric = factor(
      metric, levels = c("duration", "volume"),
      labels = c("Real duration\n(in days)", 
                 "Absolute volume\n(in 1000 m³)"))
  ) %>%
  dplyr::select(metric, quafun) %>%
  unnest(quafun)

rp.breaks <- c(1:8, seq(10, 80, 10), 100, 150, 200)

ggplot(o, aes(x = reduced_var(prob.annual), y = value)) + 
  geom_vline(xintercept = reduced_var(1 - 1/breaks_log10_all()(c(1, 100))),
             col = "lightgrey", size = 0.1) + 
  geom_point(size = 0.5) + 
  geom_line(data = f, aes(x = reduced_var(prob.annual), y = quant), size = 0.2) + 
  scale_x_continuous(
    expand = expansion(c(0.04, 0)), 
    sec.axis = sec_axis(
      trans = function(x) 1/ (1 - (-exp(-x) + 1)), 
      breaks = c(1, seq(10, 80, 10), 100, 150, 200),
      name = "Return period (in years)")
  ) + 
  scale_y_continuous(limits = c(0, NA)) + 
  facet_wrap(vars(metric), scales = "free_y", 
             ncol = 1, strip.position = "left") +
  labs(x = expression(paste("Exponential reduced variate -log(", 1, " - F(", x, "))")),
       y = "Quantile") + 
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```


### Calculation of the T-year event
The return period of a given event is calculated following Equation 6.4.
The relationship between the drought characteristics as defined in Worked
Example 5.4 and $F(x)$ are given by the tabulated distribution functions in
Nizowka. The design value for a particular return period, i.e. the T-year
event, can be obtained from the tables for known values of $F(x)$. The
estimated 10-, 100- and 200-year drought events are shown in Table 6.2.


```{r, echo = FALSE}
fitted %>%
  mutate(
    quantiles = map2(parameter, rate, ~tibble(
      prob.annual, 
      prob.pds = annual2pds(prob.annual, .y),
      quant = quagpa(prob.pds, .x)))
  ) %>%
  dplyr::select(metric, quantiles) %>%
  unnest(quantiles) %>%
  pivot_wider(id_cols = c(prob.annual, prob.pds), 
              names_from = "metric", values_from = "quant") %>%
  mutate(rp.annual = 1 / (1 - prob.annual), rp.pds = 1 / (1 - prob.pds)) %>%
  dplyr::select(prob.annual, rp.annual, rp.pds, prob.pds, duration, volume)
```

### Manual computation 

Tobias: pearhaps mention that one should not forget to set the lower bound of the GPA (location parameter) equal to the threshold.

```{r}
(parameter <- pelgpa(samlmu(pooled$duration), bound = 5))

prob.annual <- 1 - 1/ c(10, 100, 200)
prob.pds <- annual2pds(prob.annual, exc.per.year = 100 / 49)

# Quantiles of the GPA 
quagpa(f = prob.pds, para = parameter)
```






